---
title: "Black Box as a Diagnostic Reasoning Benchmark for LLMs"
subtitle: "Assessing Hidden State Identification Through Test Selection, Constraint Tracking, and Belief Updating"
author:
  - name: "Todd R. Johnson"
    affiliation: "University of Texas Health Science Center at Houston"
    email: "todd.r.johnson@uth.tmc.edu"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
bibliography: references.bib
execute:
  echo: true
  warning: false
abstract: |
  Diagnostic reasoning—identifying hidden states from observable evidence under a known causal model—is fundamental to medical diagnosis, fault detection, and troubleshooting. This study uses the classic Black Box game as a benchmark to assess LLM capabilities in diagnostic reasoning. Black Box requires the core processes of diagnosis: generating hypotheses about hidden states, selecting informative tests, tracking constraints as evidence accumulates, and updating beliefs appropriately. Unlike medical benchmarks that may reward pattern matching to training data, Black Box provides out-of-distribution problems with verifiable ground truth. We present baseline results comparing Claude model variants (Haiku, Sonnet, Opus) across prompt conditions. Our findings reveal systematic limitations in the component processes of diagnostic reasoning that persist despite scaling and prompt engineering, with implications for the deployment of LLMs in medical diagnosis and other domains requiring hidden state identification.
keywords: [large language models, diagnostic reasoning, hidden state identification, medical diagnosis, benchmark]
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(jsonlite)
library(car)
library(knitr)
library(kableExtra)
library(effectsize)
library(performance)
library(broom)
library(lme4)        # Mixed-effects models
library(lmerTest)    # p-values for mixed models
library(broom.mixed) # Tidy methods for mixed models

# Set plotting defaults
theme_set(theme_minimal())
```

# Introduction

## The Deployment Reality

AI systems are increasingly deployed for diagnostic tasks—identifying
hidden states or conditions from observable evidence. Medical AI tools
assist with diagnosis from imaging, lab results, and symptoms.
Industrial systems perform fault diagnosis on complex machinery.
Customer service agents troubleshoot technical problems through
iterative questioning.

Yet deployment has outpaced our understanding of whether these systems
actually perform diagnostic *reasoning* or sophisticated pattern
matching. @wen2025context demonstrate that high performance on medical
benchmarks—the multiple-choice question answering (MCQA) format derived
from clinical licensing examinations—may reflect "context matching"
rather than clinical reasoning. Their analysis invalidates three
assumptions underlying MCQA-based assessment: that knowledge is applied
rather than memorized, that semantic consistency leads to consistent
answers, and that models can recognize situations with no valid answer.

If LLMs achieve high scores on medical board examinations through
pattern matching rather than inference, benchmark performance provides
false assurance about deployment readiness for genuine diagnostic tasks.

## The Evaluation Crisis

The fundamental problem is that current benchmarks may not test
diagnostic reasoning at all. Most medical AI benchmarks use static
question-answering formats: given a case description, select the correct
diagnosis from options. This tests recognition, not the iterative
process of gathering evidence and refining hypotheses that characterizes
actual diagnosis.

We need benchmarks that test the *process* of diagnostic reasoning, not
just its end product. But creating such benchmarks in medical or
industrial domains is difficult: cases are complex, ground truth is
often uncertain, and training data contamination is hard to control. We
need controlled environments where the component processes of diagnostic
reasoning can be isolated and measured.

## What is Diagnostic Reasoning?

Diagnostic reasoning is the cognitive process of identifying which
hidden state (or states) is present based on observable evidence, given
a known or partially known model of how hidden states produce
observations.

**The structure of diagnostic reasoning:**

-   **Given**: A causal model mapping hidden states to observations
    (diseases cause symptoms; faults cause error codes; atom
    configurations cause ray behaviors)
-   **Task**: Identify which hidden state is present based on
    observations
-   **Process**: Iteratively gather evidence and refine hypotheses until
    confident

**Diagnostic reasoning is distinct from:**

-   **Causal structure discovery**: Learning *what causes what* (the
    causal graph itself)
-   **Causal effect estimation**: Measuring *how much* an intervention
    affects an outcome
-   **Mechanism discovery**: Understanding *how* a cause produces its
    effect

In diagnostic reasoning, we typically know (or assume) the causal
structure—we know that certain diseases cause certain symptoms, or that
certain faults produce certain error patterns. The challenge is
identifying *which* specific state from a known set is present.

**Component processes of diagnostic reasoning:**

| Process | Description |
|----|----|
| **Hypothesis Generation** | Generate candidate hidden states that could explain observations |
| **Forward Reasoning** | Predict what observations would occur under each hypothesis |
| **Test Selection** | Choose which observation to gather next to best discriminate among hypotheses |
| **Constraint Tracking** | Maintain which hidden states have been ruled in or out |
| **Belief Updating** | Revise hypothesis probabilities when new evidence arrives |
| **Stopping Decision** | Decide when evidence is sufficient to commit to a diagnosis |

These processes must be *integrated*: test selection depends on current
hypotheses; hypotheses depend on accumulated evidence; the decision to
stop depends on confidence in the leading hypothesis. Failures in any
component can cascade through the diagnostic process.

## Black Box as a Diagnostic Instrument

We propose the Black Box game as a diagnostic instrument for
understanding LLM reasoning limitations. Black Box is a classic
deduction game designed by Eric Solomon in 1978 [@blackbox1978]. Players
face an 8×8 grid containing four hidden "atoms" at unknown locations. By
firing rays into the grid and observing their behavior—absorption,
deflection, or reflection—players must infer the atom positions.

Black Box is not merely another game benchmark. It is a *minimal
realization of diagnostic reasoning*—requiring all six component
processes in an integrated task:

| Diagnostic Process | Black Box Analog |
|----|----|
| **Hypothesis Generation** | Consider which atom configurations could explain observations |
| **Forward Reasoning** | Predict ray behavior under a hypothesized configuration |
| **Test Selection** | Choose which ray to fire next to discriminate among hypotheses |
| **Constraint Tracking** | Maintain which cells are ruled out by accumulated evidence |
| **Belief Updating** | Revise candidate configurations when new ray results arrive |
| **Stopping Decision** | Decide: fire more rays or commit to a guess? |

This structural correspondence makes Black Box findings informative
about diagnostic reasoning generally, while providing advantages that
clinical or scientific domains lack:

1.  **Out-of-distribution**: While the rules are simple and fully
    specifiable, any particular game configuration is almost certainly
    not in training data
2.  **Ground truth**: We know the correct atom positions and can
    precisely measure both forward reasoning accuracy (Predict mode) and
    inverse reasoning accuracy (Play mode)
3.  **Decomposable errors**: We can analyze extended thinking traces to
    categorize failures as occurring in experiment design, world model
    simulation, constraint tracking, or belief updating

## Addressing the Spatial Reasoning Confound

A methodological concern is whether Black Box's spatial reasoning
requirements confound interpretation. Many diagnostic reasoning
problems—medical diagnosis, fault detection, troubleshooting—do not
involve spatial relationships, so spatial failures on Black Box might
not predict failures on non-spatial diagnostic tasks.

We address this through experimental design. Our two-mode structure
allows partial decomposition:

-   **Predict mode** tests forward reasoning (heavily spatial): given
    atom positions, trace the ray's path
-   **Play mode** tests the full diagnostic loop: spatial reasoning is
    one input, but experiment design, abductive inference, and
    constraint tracking are the primary challenges

Comparing performance across modes provides diagnostic information: good
Predict performance with poor Play performance would implicate
non-spatial reasoning limitations (abduction, constraint tracking,
experiment design), while poor performance on both would suggest spatial
reasoning is at least partly responsible.

Additionally, we use *mean atoms per ray*—a measure of spatial
complexity—as a moderator variable. If treatment effects (model, prompt
style, extended thinking) are constant across spatial complexity, the
limitations lie in inferential machinery rather than spatial processing.
If effects are moderated by spatial complexity, spatial reasoning
contributes to observed failures.

Finally, we analyze extended thinking traces to categorize errors:

-   **Spatial errors**: Ray path traced incorrectly
-   **Constraint errors**: Failed to track ruled-out positions
-   **Experiment design errors**: Fired uninformative rays
-   **Belief updating errors**: Did not revise hypotheses when evidence
    contradicted them

Black Box alone cannot fully separate spatial from non-spatial
limitations. However, our findings can be triangulated with non-spatial
benchmarks (ZebraLogic for constraint satisfaction, LogicGame for
rule-based planning) to identify common factors. If LLMs fail on Black
Box *and* on non-spatial reasoning benchmarks, the shared limitation is
inferential machinery, not spatial processing specifically.

## Research Questions

This study addresses the following questions:

1.  **Diagnostic Reasoning Capability**: Can LLMs perform the integrated
    reasoning loop required for diagnostic inference on novel problems?

2.  **Forward vs. Inverse Reasoning**: Do limitations arise primarily in
    world model simulation (Predict mode) or in the inferential
    machinery of hypothesis generation and updating (Play mode)?

3.  **Decomposition of Failures**: What proportion of errors reflect
    spatial reasoning failures versus non-spatial failures (constraint
    tracking, experiment design, belief updating)?

4.  **Prompt and Scaffold Effects**: Do augmented instructions,
    visualizations, or extended thinking improve performance, and do
    these effects interact with spatial complexity?

5.  **Scaling Effects**: Do larger models show improved diagnostic
    reasoning, and if so, is improvement sufficient to approach
    human-level competence?

6.  **Implications for Deployment**: What do the observed limitations
    suggest about the readiness of LLMs for medical diagnosis, fault
    detection, and other diagnostic tasks?

# Literature Review

The case for Black Box as a diagnostic instrument rests on documented
limitations in each component capability required for diagnostic
reasoning. If LLMs struggle with constraint satisfaction, spatial
reasoning, world model simulation, and abductive inference *in
isolation*, we should expect compounding failures when these
capabilities must be integrated. The literature reviewed below provides
this supporting evidence, organized by the component capabilities Black
Box requires.

A comprehensive survey by @giadikiaroglou2024puzzlesurvey provides a
useful taxonomy for understanding LLM puzzle-solving capabilities. They
distinguish between *rule-based puzzles* (with explicit victory
conditions and game rules, including deterministic puzzles like Sudoku
and Chess, and stochastic puzzles like Minesweeper and Poker) and
*rule-less puzzles* (requiring world knowledge and inference, such as
riddles and commonsense reasoning). Black Box falls squarely in the
rule-based deterministic category, making findings from this literature
directly relevant. The survey notes that advanced prompting methods like
Tree-of-Thoughts and Everything-of-Thoughts yield improvements of 26-70%
over basic Chain-of-Thought on deterministic puzzles, yet "significant
challenges in complex puzzle scenarios" persist, revealing gaps between
LLM capabilities and human-like reasoning.

## Constraint Satisfaction and LLMs

Constraint satisfaction problems (CSPs) require finding values for
variables that satisfy a set of constraints. Classic examples include
Sudoku, scheduling problems, and logic puzzles. Recent research has
systematically evaluated LLM performance on such tasks.

The ZebraLogic benchmark [@lin2024zebralogic] evaluates LLMs on logic
grid puzzles (also known as Zebra puzzles), a well-established CSP
format used in standardized tests like the LSAT. Results indicate that
while LLMs show some capability on simple puzzles, performance degrades
substantially as problem complexity increases. The benchmark reveals
that LLMs lack several abilities required for complex logical reasoning,
including counterfactual thinking, reflective reasoning, and
compositional generalization.

@bonlarron2024gencp demonstrate that LLMs "excel at generating fluent
text but struggle to enforce external constraints because they generate
tokens sequentially without explicit control mechanisms." Their GenCP
framework addresses this by combining LLM predictions with constraint
programming (CP) reasoning, treating text generation as a CSP. This
hybrid approach significantly outperforms pure LLM methods, suggesting
that constraint satisfaction may require symbolic reasoning components
that LLMs lack.

@michailidis2024constraint evaluated LLMs on translating natural
language problem descriptions into formal constraint models. They found
that zero-shot prompting yields essentially no success, and even
few-shot prompting produces inconsistent results. This aligns with
@freuder2024conversational's observation that successful constraint
satisfaction often requires iterative refinement through dialogue, not
single-shot generation.

The LogicGame benchmark [@gui2024logicgame] directly evaluates
rule-based reasoning—the ability to understand, execute, and plan
according to predefined rules. Using game scenarios at varying
difficulty levels, they found that even the best-performing model
(claude-3.5-sonnet) achieves only 19.15% overall accuracy, with
performance falling below 20% on complex tasks requiring multi-step
planning. Critically, they observed that while execution tasks improve
modestly with few-shot examples, planning tasks often *deteriorate* with
additional demonstrations, and any improvements diminish substantially
at higher difficulty levels. This suggests that LLMs struggle not just
with constraint tracking but with the multi-step planning required to
design informative experiments.

The implications for Black Box are direct: tracking which cell locations
are ruled out by each ray observation is fundamentally a constraint
satisfaction problem. If LLMs struggle with CSPs generally, they should
struggle with the constraint-tracking aspect of Black Box.

## Spatial Reasoning Limitations

Spatial reasoning—the ability to mentally represent and manipulate
spatial relationships—is a well-documented weakness of current LLMs.

The PLUGH benchmark [@tikhonov2024plugh] evaluates spatial understanding
and reasoning using text adventure game scenarios that require building
mental maps from textual descriptions. Results show that while some
commercial LLMs exhibit moderate reasoning abilities, "all models still
have significant room for improvement." The benchmark identifies typical
failure modes including difficulty maintaining consistent spatial
representations across multiple textual updates.

@chen2024spatial conducted an empirical analysis of spatial reasoning in
large multimodal models, finding that even GPT-4o achieves only 27.5%
accuracy on questions requiring reasoning from a human's viewpoint in an
image. Performance degrades further on multi-hop spatial reasoning
questions that require integrating multiple spatial relationships.

@bos2024spatial provides a practitioner's perspective on LLM spatial
reasoning, noting that "all of the LLMs failed immediately on the
easiest problems" involving mental box folding—a standard spatial
reasoning task in human cognitive assessments. Interestingly, LLMs
performed better on reasoning about objects in photographs than on
abstract spatial diagrams, suggesting reliance on statistical patterns
from image-caption training data rather than genuine spatial reasoning.

A synthesis of spatial reasoning research [@quan2025stark] notes that
"empirical assessments consistently show a characteristic pattern:
models exhibit moderate competence in simple, direct spatial tasks but
rapidly deteriorate as the problem scale or compositional/geometric
complexity increases, with performance losses ranging from 42% to over
80% as task complexity grows."

The BALROG benchmark [@paglieri2024balrog] evaluates LLMs and vision
language models on diverse game environments ranging from tasks
"solvable by non-experts in seconds to extremely difficult challenges
like NetHack requiring years to master." Their findings reveal that
current models achieve only partial success on simpler games but
"struggle significantly with more challenging tasks." Most striking is
their discovery of "severe deficiencies in vision-based
decision-making"—several models actually perform *worse* when given
visual representations of environments compared to text-only
descriptions. This counterintuitive result suggests that visual
processing may introduce noise rather than helpful spatial information.

@wang2025crosswordbench addresses a related challenge: tasks requiring
both semantic understanding and adherence to visual grid constraints.
Using crossword puzzles as a benchmark, they demonstrate that
"mathematical reasoning training doesn't necessarily generalize to tasks
requiring linguistic and spatial integration." This finding is
particularly relevant to Black Box, which similarly requires integrating
rule-based reasoning (ray physics) with spatial constraint tracking
(grid positions).

For Black Box, ray tracing requires understanding how rays move through
2D space and how they deflect upon encountering atoms. The documented
limitations in spatial reasoning suggest LLMs will struggle with
accurate ray path simulation.

## Reasoning Complexity and Performance Collapse

A critical question for evaluating LLM reasoning is how performance
scales with problem complexity. @shojaee2025illusion provide a
systematic investigation using controllable puzzle environments that
allow precise manipulation of compositional complexity while maintaining
consistent logical structures.

Their findings reveal three distinct performance regimes. In
low-complexity tasks, standard models surprisingly outperform reasoning
models (those using extended thinking or chain-of-thought), suggesting
that extended reasoning can introduce unnecessary complexity for simple
problems. In medium-complexity tasks, extended thinking provides clear
advantages, demonstrating the intended benefit of reasoning-augmented
models. However, in high-complexity tasks, both standard and reasoning
models exhibit complete accuracy collapse—a fundamental ceiling that
persists regardless of model capability or available computational
resources.

Particularly striking is their observation that reasoning models exhibit
*declining effort* at higher complexity levels despite adequate token
budgets. Rather than leveraging available computational resources more
intensively when problems become harder, models show inconsistent
reasoning patterns across similar problems. This suggests that current
reasoning mechanisms are not genuinely algorithmic but rather
approximate heuristics that fail when problems exceed certain complexity
thresholds.

The PuzzlePlex benchmark [@puzzleplex2025] extends this analysis to
multi-turn adversarial environments, evaluating LLM reasoning and
planning across 24 diverse puzzles including both deterministic and
stochastic games. Their findings are sobering: "LLMs still significantly
lag behind even simple heuristics in puzzles." This result suggests that
the limitations are not merely about single-step reasoning accuracy but
extend to the sequential decision-making and strategic planning that
characterizes game play. For tasks requiring iterative hypothesis
refinement based on accumulated evidence—like Black Box—this predicts
systematic failures in experiment design and belief updating.

For Black Box, these findings predict a non-linear relationship between
game difficulty (determined by atom configurations, number of atoms, and
observation ambiguity) and model performance. Simple configurations may
be solved adequately, but complex configurations requiring deep
inference chains should trigger the same "reasoning collapse" observed
across other puzzle domains.

## World Models and Mental Simulation

A "world model" refers to an internal representation that can be used to
simulate how states evolve over time. The question of whether LLMs
possess genuine world models—versus sophisticated pattern matching—is
actively debated.

@harig2025worldmodels investigated this question using mechanical
reasoning about pulley systems. Their findings are striking: while LLMs
showed some ability to estimate mechanical advantage, they appeared to
use a "pulley counting heuristic" rather than genuine simulation. The
best model (GPT-4o) achieved only 26.1% accuracy on exact predictions.
When problems were modified to be superficially similar but mechanically
different, performance collapsed, suggesting "abilities are often
fragile and dependent on superficial features rather than deep
understanding."

@wang2024bytesized directly tested whether LLMs can serve as text-based
world simulators. Using a benchmark of 76,369 state transitions from
text adventure games, they found that GPT-4's best single-step accuracy
was 59.9%. Because simulation errors accumulate, after 10 steps the
expected accuracy drops to less than 1% ($0.599^{10}$). They conclude
that "LLMs are not yet able to reliably act as text world simulators."

This literature suggests that LLMs may appear to reason about world
dynamics but are actually relying on heuristic shortcuts that fail when
problems require genuine simulation. For Black Box's "Predict"
mode—where the model must trace a ray's path given known atom
positions—we should expect systematic errors reflecting heuristic
approximation rather than accurate simulation.

However, recent work suggests that targeted post-training on
decision-making games can substantially improve LLM reasoning in those
specific domains. @wang2025mastermind developed data synthesis
strategies for training LLMs on Doudizhu (a Chinese card game) and Go,
demonstrating that game-based training data can complement traditional
code and mathematics datasets for reasoning enhancement. Their
"Mastermind" models achieved 90% action accuracy matching expert data
and 98% similarity in thought processes for Doudizhu. Critically, this
approach uses hierarchical training pipelines that decompose complex
decisions into subtasks—possible action prediction, opponent strategy
modeling, and final action selection—rather than relying on direct
policy learning.

For our purposes, this work highlights an important distinction: Black
Box is designed as an *out-of-distribution* benchmark that tests
zero-shot reasoning capabilities on novel configurations. The success of
task-specific training reported by @wang2025mastermind does not
contradict predictions of poor zero-shot performance; rather, it
suggests that observed limitations may reflect training distribution
gaps rather than fundamental architectural constraints. Whether
fine-tuning on Black Box traces would improve performance—and whether
such improvement would generalize to other spatial reasoning
tasks—remains an empirical question addressed in our future directions.

## Abductive Reasoning and Hypothesis Generation

Abductive reasoning—inferring the most plausible explanation for
observations—is the core inferential process in diagnostic reasoning. A
growing body of research examines LLM capabilities in this domain.

@yang2025survey provides a comprehensive framework for understanding LLM
hypothesis discovery, organizing methods according to Peirce's reasoning
paradigm: abduction (hypothesis generation), deduction (hypothesis
application), and induction (hypothesis validation). They note that
"most existing work assessing LLM reasoning centers almost exclusively
on multi-step deductive tasks, leaving abduction and induction, the
engines of hypothesis discovery, largely unexplored."

@wang2025occam directly tests whether LLMs follow Occam's Razor—the
principle of preferring simpler explanations. Using a synthetic
benchmark (InAbHyD) designed to test inductive and abductive reasoning,
they find that "current LLMs can't generate high-quality hypotheses that
match human-level intelligence." LLMs frequently produce overly complex
explanations when simpler ones suffice, failing a fundamental principle
of diagnostic reasoning.

@liu2024incomplete characterizes the current state as "an incomplete
loop," arguing that while LLMs can perform some aspects of each
reasoning type, they fail to integrate abduction, deduction, and
induction into a coherent reasoning cycle. This fragmentation prevents
genuine hypothesis-driven inquiry.

@rodriguez2025analogical argues from a theoretical perspective that LLMs
face fundamental limitations in abductive reasoning: "(1) They cannot
work with counterfactuals, relying only on correlational datasets. (2)
They are unable to perform true abductive reasoning." While LLMs may
appear to generate hypotheses, this reflects parameter-controlled
variation rather than genuine inference to the best explanation.

For Black Box, the inverse problem—inferring atom locations from ray
observations—is quintessentially abductive. The literature predicts
systematic failures in generating parsimonious hypotheses and in
appropriately updating beliefs as evidence accumulates.

## Test Selection in Diagnostic Reasoning

A distinguishing feature of diagnostic reasoning is *active* test
selection—choosing which observations to gather based on current
hypotheses to maximally discriminate among candidate diagnoses. Most LLM
benchmarks present static problems where all information is provided
upfront. Few benchmarks test the ability to strategically select tests
and iteratively refine hypotheses.

Auto-Bench [@autobench2025] evaluates LLMs on a task structurally
similar to diagnosis: discovering hidden graph structures through
iterative experimentation. The task proceeds in cycles: the model
receives previous observations, proposes a hypothesis about the hidden
structure, and designs an intervention to test that hypothesis. An
oracle executes the intervention and returns new observations,
continuing until the hypothesis matches ground truth or a limit is
reached. This directly parallels diagnostic reasoning: observe effects,
hypothesize hidden states, select tests to discriminate. The key finding
is that LLMs struggle with strategic test selection—they often propose
uninformative experiments that fail to disambiguate among hypotheses.

This finding is particularly relevant for Black Box. Effective play
requires selecting rays that maximally constrain the hypothesis space—
firing into regions where different candidate configurations would
produce different outcomes. Random or poorly-chosen rays waste resources
and may never converge on the correct answer. The documented failures in
Auto-Bench predict that LLMs will struggle with strategic ray selection
in Black Box's Play mode.

## Base Rate Neglect in Clinical Diagnosis

A particularly consequential manifestation of abductive reasoning
failure is base rate neglect—the tendency to overweight salient
observations while underweighting prior probabilities. @omar2025zebras
provide the first large-scale demonstration of this phenomenon in LLMs
applied to clinical diagnosis.

Using 300 synthetic clinical vignettes tested across 20 LLMs with
approximately 1.8 million responses, they found that models
systematically selected rare "zebra" diagnoses linked to salient but
non-decisive cues (such as travel history or animal exposure) more often
than epidemiologically correct diagnoses (49.8% vs 37.7%). This pattern
persisted even when models were explicitly asked to identify the "most
likely" diagnosis. Critically, prompts that explicitly invoked base
rates and epidemiology reduced but did not eliminate this saliency bias.

The implications are profound: LLMs that achieve high performance on
medical board examinations—demonstrating factual knowledge of disease
prevalence and diagnostic criteria—nevertheless fail to appropriately
weight this knowledge when salient distractors are present. As the
authors note, this represents "associative pattern matching rather than
stable probabilistic reasoning." Training on medical literature, which
overrepresents rare and dramatic conditions in case reports and teaching
materials, may systematically bias models toward zebra diagnoses.

For Black Box, this finding suggests that even when LLMs possess
accurate knowledge of ray physics, they may be systematically distracted
by salient features of observations (e.g., absorption events, unusual
deflection patterns) at the expense of systematic constraint tracking
and hypothesis evaluation. The base rate neglect documented in clinical
diagnosis may manifest as premature hypothesis commitment or failure to
consider the full space of configurations consistent with accumulated
evidence.

## Implications for Diagnostic AI Systems

The limitations documented above converge on a concerning picture for
LLM deployment in diagnostic contexts.

The LLM-Modulo framework [@kambhampati2024llmmodulo;
@valmeekam2023planning] proposes a pragmatic response: accept that LLMs
cannot plan or reason reliably, but use them as components within hybrid
systems that include external verifiers and symbolic reasoners. Under
this view, the LLM's role is translation and suggestion, not autonomous
diagnostic reasoning. A physician might use an LLM to suggest
differential diagnoses, but should not rely on it to track constraints
or select optimal tests.

This matters because AI diagnostic systems are already being deployed.
Medical AI tools assist with diagnosis from imaging, lab results, and
symptoms. Industrial systems perform fault diagnosis on complex
machinery. The failure modes documented throughout this review—failure
to track constraints, inability to select informative tests, base rate
neglect, and premature hypothesis commitment—map directly onto the
requirements of diagnostic reasoning.

Diagnostic benchmarks that can characterize *when* and *why* these
failures occur are needed before broader deployment in high-stakes
diagnostic domains.

## Summary: Predictions for Diagnostic Reasoning in Black Box

Based on the documented limitations in component processes of diagnostic
reasoning, we derive the following predictions:

1.  **Forward reasoning (Predict mode)** will show substantial errors,
    consistent with documented failures in spatial reasoning and world
    model simulation.

2.  **Inverse reasoning (Play mode)** will show even greater errors, as
    it compounds spatial reasoning with abductive inference and
    constraint satisfaction.

3.  **Prompt engineering** will provide modest improvements but will not
    overcome fundamental limitations.

4.  **Scaling** (larger models) will show improvement but not to
    human-level performance.

5.  **Error patterns** will reveal heuristic approximation rather than
    systematic reasoning—e.g., over-reliance on surface features,
    failure to track constraints, non-parsimonious hypotheses.

6.  **Tool augmentation** (e.g., giving the LLM a ray-tracing verifier)
    will help with verification but not with hypothesis generation, as
    the limitation is in search/inference, not just simulation.

7.  **Uninformative test selection**: Following findings from Auto-Bench
    [@autobench2025], models will frequently choose rays that fail to
    disambiguate among hypotheses—firing into already-constrained
    regions or missing opportunities for informative observations.
    Strategic test selection requires anticipating which observations
    would most reduce uncertainty, a capability that current models
    lack.

8.  **Complexity-dependent collapse**: Following @shojaee2025illusion,
    performance will show a non-linear pattern—adequate performance on
    simple configurations but catastrophic failure on complex
    configurations requiring extended inference chains, regardless of
    available thinking budget.

9.  **Saliency-driven distraction**: Following @omar2025zebras, models
    will be systematically distracted by salient observations (e.g.,
    absorption events, dramatic deflections) and may anchor prematurely
    on hypotheses suggested by striking but non-decisive evidence, even
    when simpler explanations are more consistent with the full
    observation set.

10. **Spatial-inferential decomposition**: If spatial reasoning is the
    primary bottleneck, we expect: (a) poor Predict mode performance,

    (b) treatment effects moderated by spatial complexity (mean atoms
        per ray), and (c) error traces dominated by ray path
        miscalculations. If inferential machinery is the primary
        bottleneck, we expect:
    (c) relatively better Predict than Play performance, (b) treatment
        effects constant across spatial complexity, and (c) error traces
        showing constraint tracking failures, uninformative experiment
        choices, and belief updating errors.

11. **Convergent validity**: If Black Box limitations reflect general
    diagnostic reasoning failures rather than spatial-specific failures,
    we expect correlations with findings from non-spatial benchmarks
    (ZebraLogic, LogicGame). Models that perform poorly on constraint
    satisfaction and rule-based planning should also perform poorly on
    Black Box's non-spatial components.

# Methods

## The Black Box Game Implementation

We implemented Black Box as a web-based application using React,
allowing both human play and automated LLM interaction through the
Anthropic API. The implementation follows standard Black Box rules:

-   **Grid**: 8×8 cells, with 4 randomly placed atoms (non-adjacent
    placement enforced)
-   **Rays**: Fired from any of 32 edge positions (8 per side)
-   **Ray behavior**:
    -   Direct hit on atom: **Absorbed** (marked "H" for hit)
    -   Atom diagonally adjacent to entry cell: **Reflected** (marked
        "R")
    -   Atom diagonally adjacent to path: **Deflected** 90° away from
        atom
    -   Atoms on both sides of path: **Reflected** (reverses direction)
    -   Clear path: Ray exits at opposite edge

## Task Modes

### Predict Mode (Forward Reasoning)

In Predict mode, the LLM is shown: - The 8×8 grid with atom positions
marked - A ray entry point (e.g., "NORTH-4")

The LLM must predict the ray's outcome: exit position, absorption, or
reflection. This tests world model accuracy—can the LLM correctly
simulate ray physics?

### Play Mode (Inverse Reasoning)

In Play mode, the LLM: - Receives the game rules - Chooses which rays to
fire (up to 20) - Receives feedback after each ray (exit position,
absorption, or reflection) - Must guess all 4 atom positions

This tests the full reasoning pipeline: experiment design, observation
integration, constraint tracking, and abductive inference.

## Prompt Conditions

We developed two prompt styles with optional visualization augmentation:

### Baseline (Human-Equivalent)

Comprehensive instructions adapted from the traditional Emacs Black Box
rules, comparable to what a human player would receive. The prompt
includes:

-   Complete rules with ASCII diagram examples showing deflection,
    reflection, and absorption
-   Scoring information (rays cost points, missed atoms cost 5 points
    each)
-   JSON response format specification

Key excerpt:

```         
Your opponent has hidden 4 balls within this box. By shooting rays into
the box and observing where they emerge, it is possible to deduce the
positions of the hidden balls.

There are three possible outcomes for each ray you send into the box:
- DETOUR: The ray is deflected and emerges somewhere other than where
  you sent it in.
- REFLECTION (R): The ray is reflected and emerges in the same place
  it was sent in.
- HIT (H): The ray strikes a ball directly and is absorbed.
```

### Augmented

Extended instructions including:

-   Explicit coordinate system explanation
-   Detailed deflection geometry with worked examples
-   Strategy guidance (triangulation, multiple rays)
-   Common error warnings
-   Explicit note that single observations are typically ambiguous

### Visualization Options

Either prompt style can be combined with:

-   **Include Visualization**: Text-based board visualization showing
    current state
-   **VoT (Visualization of Thought) Options**:
    -   Grid State (Play only): Shows current board state in prompt
    -   Ray Trace: Shows ray path visualization
    -   Hypothesis (Play only): Shows marked hypothesis positions

## Models

We evaluated three Claude 4.5 model variants:

-   **Claude Haiku 4.5** (`claude-haiku-4-5-20251001`): Smallest,
    fastest
-   **Claude Sonnet 4.5** (`claude-sonnet-4-5-20250929`): Mid-tier
-   **Claude Opus 4.5** (`claude-opus-4-5-20251101`): Largest, most
    capable

All models were accessed through the Anthropic API with extended
thinking enabled to capture reasoning processes.

## Experimental Design

The experiment uses a factorial design with factors that differ slightly
between Predict and Play modes. Ten fixed atom configurations ensure
reproducibility across conditions.

### Predict Mode Factors

```{r}
#| label: predict-design-table
#| echo: false

predict_design <- tibble(
  Factor = c('Model', 'Prompt Style', 'Extended Thinking', 'VoT Ray Trace', 'Atom Configuration'),
  Levels = c('Haiku 4.5, Sonnet 4.5, Opus 4.5',
             'Baseline, Augmented',
             'Yes (with budget), No',
             'Yes, No',
             '10 fixed configurations'),
  Description = c('Model variant',
                  'Human-equivalent vs. detailed strategy guidance',
                  'Enable extended thinking with token budget',
                  'Include ray trace visualization in prompt',
                  'Predetermined atom placements for reproducibility')
)

kable(predict_design, format = "pipe")
```

| Model | Prompt | Extended Thinking | Vot (None, B Ray Trace) |
|----|----|----|----|
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **No** | **None** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **No** | **VoT B Ray Trace Visualization** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **10K** | **None** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **10K** | **VoT B Ray Trace Visualization** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **No** | **None** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **No** | **VoT B Ray Trace Visualization** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **10K** | **None** |
| **\*Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **10K** | **VoT B Ray Trace Visualization** |

### Play Mode Factors

Play Mode Experiment Log

| Model | Prompt | Extended Thinking | VoT (None, A, B, or C) |
|----|----|----|----|
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **No** | **None** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **No** | **VoT A Grid State Tracking** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **No** | **VoT B Ray Trace Visualization** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **No** | **VoT C Hypothesis Verification** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **10K** | **None** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **10K** | **VoT A Grid State Tracking** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **10K** | **VoT B Ray Trace Visualization** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **baseline** | **10K** | **VoT C Hypothesis Verification** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **No** | **None** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **No** | **VoT A Grid State Tracking** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **No** | **VoT B Ray Trace Visualization** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **No** | **VoT C Ray Trace Visualization** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **10K** | **None** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **10K** | **VoT A Grid State Tracking** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **10K** | **VoT B Ray Trace Visualization** |
| **Haiku 4.5, Sonnet 4.5, Opus 4.5** | **augmented** | **10K** | **VoT B Ray Trace Visualization** |

: Play Mode Experiment Log

Play mode includes all Predict mode factors plus additional factors for
hypothesis management:

```{r}
#| label: play-design-table
#| echo: false

play_design <- tibble(
  Factor = c('Model', 'Prompt Style', 'Include Visualization', 'Extended Thinking', 'Allow Hypotheses', 'VoT Grid State', 'VoT Ray Trace', 'VoT Hypothesis', 'Atom Configuration'),
  Levels = c('Haiku 4.5, Sonnet 4.5, Opus 4.5',
             'Baseline, Augmented',
             'Yes, No',
             'Yes (with budget), No',
             'Yes, No',
             'Yes, No',
             'Yes, No',
             'Yes, No',
             '10 fixed configurations'),
  Description = c('Model variant',
                  'Human-equivalent vs. detailed strategy guidance',
                  'Text board visualization in prompt',
                  'Enable extended thinking with token budget',
                  'Enable mark/unmark/check actions for hypothesis tracking',
                  'Show current grid state in prompt (VoT Option A)',
                  'Include ray trace visualization (VoT Option B)',
                  'Show hypothesized atom positions (VoT Option C)',
                  'Predetermined atom placements for reproducibility')
)

kable(play_design, format = "pipe")
```

### Atom Configurations

Ten fixed atom configurations are used for reproducible experiments:

```{r}
#| label: config-table
#| echo: false

configs <- tibble(
  Config = 1:10,
  Pattern = c('Spread', 'Cluster in corner', 'Diagonal', 'Edge-heavy', 'Central cluster',
              'L-shape', 'Corners', 'Asymmetric', 'Row cluster', 'Mixed'),
  Positions = c('(2,3), (3,6), (6,2), (7,7)',
                '(1,1), (1,3), (2,2), (5,6)',
                '(2,2), (4,4), (6,6), (8,8)',
                '(1,4), (4,8), (8,5), (5,1)',
                '(3,4), (4,3), (4,5), (5,4)',
                '(2,2), (2,3), (2,4), (4,2)',
                '(1,1), (1,8), (8,1), (8,8)',
                '(2,7), (3,2), (6,5), (7,3)',
                '(4,2), (4,4), (4,6), (4,8)',
                '(1,5), (3,3), (5,7), (8,2)')
)

kable(configs, format = "pipe")
```

### Predict Mode Procedure

For each experimental condition (model × prompt style × visualization ×
thinking × VoT):

1.  For each of the 10 fixed atom configurations:
    a.  For each of 32 possible ray entry points:
        -   Present atom positions and ray entry to LLM
        -   Record predicted outcome (exit position, absorption, or
            reflection)
        -   Compare to ground-truth ray trace
    b.  Compute accuracy metrics for this configuration
2.  Aggregate accuracy across configurations

### Play Mode Procedure

For each experimental condition (model × prompt style × visualization ×
thinking × hypotheses × VoT options):

1.  For each of the 10 fixed atom configurations:
    a.  Initialize game with hidden atom positions
    b.  LLM selects rays iteratively (max 20)
    c.  After each ray, provide outcome feedback
    d.  If hypotheses enabled, LLM can mark/unmark suspected positions
    e.  LLM submits final guess of 4 atom positions (via guess or check
        action)
    f.  Record: rays used, invalid moves, atoms correct, reasoning
        trace, thinking
2.  Aggregate performance metrics across configurations

## Dependent Variables

### Predict Mode

-   **Accuracy**: Proportion of rays with correct outcome prediction
-   **Error types**: Absorption/deflection/reflection confusion patterns
-   **Response time**: API response latency (ms)
-   **Token usage**: Input and output tokens per prediction

### Play Mode

-   **Atoms correct**: 0-4 atoms in correct positions
-   **Rays used**: Number of rays before guessing
-   **Invalid moves**: Attempts to fire into used positions
-   **Efficiency**: Atoms correct / rays used
-   **Score**: Game score (lower is better: ray costs + 5 × missed
    atoms)
-   **Response time**: Total API response time (ms)
-   **Token usage**: Total input and output tokens

### Process Measures (from Extended Thinking)

-   **Thinking traces**: Full extended thinking content when enabled
-   **Reasoning explanations**: Model's stated reasoning for each action
-   **Hypothesis tracking**: Marked positions over time (when enabled)
-   **Constraint tracking accuracy**: Quality of constraint maintenance
-   **Hypothesis revision patterns**: How beliefs change with new
    evidence

## Analysis Plan

### Statistical Approach

Unlike most LLM benchmarks that report only accuracy rankings, we employ
factorial ANOVA with effect size estimation to quantify the unique
contribution of each experimental factor. This approach allows us to
determine not just *which* conditions produce better performance, but
*how much* each factor contributes and whether effects depend on problem
complexity (interaction effects). While less common in LLM evaluation,
this statistical framework is standard in experimental psychology and
provides stronger inferences about the mechanisms underlying diagnostic
reasoning failures.

We complement ANOVA with regression analysis to provide interpretable
effect magnitudes:

-   **ANOVA**: Tests omnibus hypotheses (does model matter? does prompt
    style matter?) and interaction effects
-   **Regression**: Quantifies effect magnitudes (how much does Opus
    improve over Haiku?) and allows inclusion of continuous covariates
    (spatial complexity)
-   **Effect sizes**: Standardized metrics (η², Cohen's d, odds ratios)
    allow comparison across studies and DVs

### Primary Analyses

1.  **Predict Mode**: Logistic regression with Type III ANOVA on
    prediction accuracy (binary outcome)
2.  **Play Mode**: Linear regression with Type III ANOVA on atoms
    correct (continuous outcome)
3.  **Scaling analysis**: Linear trend across model sizes
4.  **Effect modification**: Test whether treatment effects vary with
    spatial complexity (mean atoms per ray)

### Effect Size Interpretation Guidelines

Following Cohen (1988), we interpret effect sizes using conventional
benchmarks:

| Measure | Small | Medium | Large | Notes |
|----|----|----|----|----|
| η² (Eta-squared) | 0.01 | 0.06 | 0.14 | Total variance explained |
| η²p (Partial eta-squared) | 0.01 | 0.06 | 0.14 | Variance explained controlling for other factors |
| ω² (Omega-squared) | 0.01 | 0.06 | 0.14 | Unbiased estimate (preferred over η²) |
| Cohen's d | 0.20 | 0.50 | 0.80 | Standardized mean difference |
| Cohen's f | 0.10 | 0.25 | 0.40 | Effect size for ANOVA (f = √\[η²/(1-η²)\]) |
| Cramér's V | 0.10 | 0.30 | 0.50 | Association strength for categorical variables |
| Odds Ratio | 1.5 | 3.0 | 9.0 | OR = 1 means no effect; \> 1 increases odds |

Note: These are heuristics, not absolute cutoffs. Practical significance
depends on context and cost-benefit considerations.

### Qualitative Analysis

Systematic coding of extended thinking traces to identify: - Constraint
tracking errors - Spatial reasoning errors - Hypothesis generation
patterns - Response to contradictory evidence

### Spatial-Inferential Decomposition Analysis

To address the concern that spatial reasoning requirements may confound
interpretation for non-spatial diagnostic tasks, we employ three
complementary analytical strategies:

**1. Mode Comparison**

Comparing Predict mode (forward reasoning, heavily spatial) to Play mode
(full diagnostic loop) performance provides diagnostic information:

-   *Good Predict + Poor Play*: Implicates non-spatial limitations
    (abductive inference, constraint tracking, experiment design)
-   *Poor Predict + Poor Play*: Implicates spatial/world model
    limitations as at least partly responsible
-   *Performance delta* (Play accuracy conditional on Predict accuracy):
    Isolates the contribution of inferential machinery

**2. Spatial Complexity Moderation**

We use two complementary measures of spatial complexity:

-   **Predict mode**: *Cells traveled*—the total number of cells a ray
    traverses (entry cell + internal path + exit cell). This measures
    path length complexity.
-   **Play mode**: *Mean atoms per ray*—the average number of atoms
    affecting each ray's path in a game. This measures interaction
    complexity.

Moderation analysis tests whether treatment effects (model, prompt
style, extended thinking) interact with spatial complexity:

-   *Constant effects across complexity*: Limitations lie in inferential
    machinery, not spatial processing
-   *Effects moderated by complexity*: Spatial reasoning contributes to
    observed failures

**3. Error Taxonomy from Thinking Traces**

We classify errors from extended thinking traces into categories:

| Error Type | Spatial? | Description |
|----|----|----|
| Ray path error | Yes | Path traced incorrectly through grid |
| Deflection geometry error | Yes | Misunderstood deflection rules |
| Constraint tracking error | No | Failed to update ruled-out positions |
| Experiment design error | No | Fired uninformative/redundant rays |
| Belief updating error | No | Did not revise hypothesis when contradicted |
| Premature commitment | No | Anchored on hypothesis before sufficient evidence |

The proportion of spatial vs. non-spatial errors indicates which
component capabilities drive overall performance.

# Results

## Predict Mode Results

```{r}
#| label: load-predict-data

# Load all Predict mode experiment JSON files from Experiment 1/Predict/
predict_json_files <- list.files(
  path = "Experiment 1/Predict",
  pattern = "\\.json$",
  full.names = TRUE
)

# Helper function to count cells traveled by a ray
# Entry cell (outside box) + internal path cells + exit cell (outside box)
count_ray_cells <- function(rayResult) {
  if (is.null(rayResult) || is.null(rayResult$path)) {
    return(NA_real_)
  }

  path <- rayResult$path
  if (!is.matrix(path) || nrow(path) == 0) {
    return(NA_real_)
  }

  # Start with entry cell (outside box) + all internal path cells
  cells <- 1 + nrow(path)

  # Add exit cell (outside box) for rays that exit (not absorbed)
  # Absorbed rays have absorbed=TRUE and no exit field
  # Deflected/reflected rays have absorbed=FALSE and an exit field
  if (!is.null(rayResult$absorbed) && !rayResult$absorbed) {
    cells <- cells + 1
  }

  return(cells)
}

# Helper function: count atoms adjacent (within 1 cell) to any cell in ray path
# Note: jsonlite parses arrays as matrices, so atom_config is 4x2 and path is Nx2
count_atoms_affecting <- function(path, atom_config) {
  if (is.null(path) || length(path) == 0) return(0)
  if (!is.matrix(path)) return(0)

  affected <- 0
  n_atoms <- nrow(atom_config)
  n_cells <- nrow(path)

  for (i in seq_len(n_atoms)) {
    atom_row <- atom_config[i, 1]
    atom_col <- atom_config[i, 2]
    for (j in seq_len(n_cells)) {
      cell_row <- path[j, 1]
      cell_col <- path[j, 2]
      if (abs(atom_row - cell_row) <= 1 && abs(atom_col - cell_col) <= 1) {
        affected <- affected + 1
        break  # Count each atom only once per ray
      }
    }
  }
  affected
}

# Function to extract results from a single Predict JSON file
extract_predict_results <- function(json_file) {
  data <- jsonlite::fromJSON(json_file, simplifyDataFrame = FALSE)

  # Extract results array
  results <- data$results

  # Convert each result to prediction-level rows
  map_dfr(results, function(r) {
    # Each result has a predictions array
    predictions <- r$predictions

    if (is.null(predictions) || length(predictions) == 0) {
      return(tibble())
    }

    # Extract each prediction
    map_dfr(predictions, function(p) {
      # Calculate path complexity metrics
      cells_traveled <- count_ray_cells(p$rayResult)
      atoms_affecting <- count_atoms_affecting(p$rayResult$path, r$atomConfig)

      tibble(
        experiment_id = r$experimentId,
        model = r$model,
        model_name = r$modelName,
        prompt_style = r$promptStyle,
        prompt_condition = r$promptCondition,
        config_index = r$configIndex,
        include_visualization = r$includeVisualization,
        enable_thinking = r$enableThinking,
        thinking_budget = r$thinkingBudget,
        vot_ray_trace = r$votRayTrace,
        ray_entry = p$rayEntry,
        predicted = p$predicted,
        actual = p$actual,
        correct = p$correct,
        reasoning = p$reasoning,
        response_time_ms = p$responseTimeMs,
        input_tokens = p$inputTokens,
        output_tokens = p$outputTokens,
        cells_traveled = cells_traveled,
        atoms_affecting = atoms_affecting,
        source_file = basename(json_file)
      )
    })
  })
}

# Load all results into a single long-format tibble
predict_data <- map_dfr(predict_json_files, extract_predict_results)
```

Data loaded: `r nrow(predict_data)` predictions from `r length(predict_json_files)` JSON files.

- **Models**: `r paste(unique(predict_data$model_name), collapse = ", ")`
- **Prompt styles**: `r paste(unique(predict_data$prompt_style), collapse = ", ")`
- **Extended thinking**: `r paste(unique(predict_data$enable_thinking), collapse = ", ")`
- **VoT ray trace**: `r paste(unique(predict_data$vot_ray_trace), collapse = ", ")`
- **Configurations tested**: `r length(unique(predict_data$config_index))`
- **Cells traveled range**: `r round(min(predict_data$cells_traveled, na.rm = TRUE), 2)`–`r round(max(predict_data$cells_traveled, na.rm = TRUE), 2)`
- **Atoms affecting range**: `r min(predict_data$atoms_affecting, na.rm = TRUE)`–`r max(predict_data$atoms_affecting, na.rm = TRUE)`

```

### Data Completeness Check

**Note on sample size**: Fewer than 32 predictions per configuration is expected due to ray symmetry optimization. When a ray enters at position A and exits at B (detour), the experiment skips testing the reverse direction to avoid redundancy.

```{r}
#| label: predict-data-completeness

predictions_per_config <- predict_data |>
  group_by(model_name, prompt_style, enable_thinking, vot_ray_trace, config_index) |>
  summarise(n_predictions = n(), .groups = "drop")

# Create matrix: rows = experimental conditions, columns = configurations
# Since all models should have same N per config, take first model's values
predictions_matrix <- predictions_per_config |>
  filter(model_name == first(model_name)) |>
  mutate(
    condition = paste0(
      substr(prompt_style, 1, 3), "_",
      ifelse(enable_thinking, "think", "no"), "_",
      ifelse(vot_ray_trace, "vot", "no")
    )
  ) |>
  select(condition, config_index, n_predictions) |>
  pivot_wider(names_from = config_index, values_from = n_predictions, names_prefix = "cfg_")

kable(predictions_matrix,
      caption = "Predictions per Configuration by Experimental Condition",
      col.names = c("Condition", paste0("Config ", 0:9)))

# Check for consistent N across models within each condition
consistency <- predictions_per_config |>
  group_by(prompt_style, enable_thinking, vot_ray_trace, config_index) |>
  summarise(
    n_unique = n_distinct(n_predictions),
    min_n = min(n_predictions),
    max_n = max(n_predictions),
    .groups = "drop"
  )

# Check for duplicate runs (same condition run multiple times)
runs_per_condition <- predict_data |>
  group_by(prompt_style, enable_thinking, vot_ray_trace, model_name, config_index) |>
  summarise(n_runs = n_distinct(source_file), .groups = "drop")
```

**Consistency**: `r if(all(consistency$n_unique == 1)) "All models have consistent N per configuration within each condition." else "Inconsistent N detected across models."`

**Duplicate runs**: `r if(any(runs_per_condition$n_runs > 1)) "Found duplicate runs in some conditions." else "No duplicate runs detected."`

```{r}
#| label: predict-results-summary

# Summary statistics by all experimental factors
predict_summary <- predict_data |>
  group_by(model_name, prompt_style, enable_thinking, vot_ray_trace) |>
  summarise(
    n = n(),
    n_correct = sum(correct),
    accuracy = mean(correct),
    accuracy_se = sqrt(accuracy * (1 - accuracy) / n),
    mean_response_time = mean(response_time_ms, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !c(n, n_correct), \(x) round(x, 3)))

kable(predict_summary,
      col.names = c("Model", "Prompt", "Thinking", "VoT Ray", "N", "Correct",
                    "Accuracy", "SE", "Response Time (ms)"),
      caption = "Predict Mode Performance by Experimental Condition")
```

### Predict Mode Leaderboard

Top 10 configurations by overall accuracy. Format matches typical LLM benchmark reporting for cross-study comparison.

```{r}
#| label: predict-leaderboard

leaderboard_predict <- predict_data |>
  mutate(
    condition = paste0(
      model_name, " | ",
      ifelse(prompt_style == "augmented", "Aug", "Base"), " | ",
      ifelse(enable_thinking, "Think", "No-Think"), " | ",
      ifelse(vot_ray_trace, "VoT", "No-VoT")
    )
  ) |>
  group_by(condition, model_name, prompt_style, enable_thinking, vot_ray_trace) |>
  summarise(
    n = n(),
    accuracy = mean(correct),
    ci_lower = accuracy - 1.96 * sqrt(accuracy * (1 - accuracy) / n),
    ci_upper = accuracy + 1.96 * sqrt(accuracy * (1 - accuracy) / n),
    .groups = "drop"
  ) |>
  arrange(desc(accuracy)) |>
  mutate(
    rank = row_number(),
    accuracy_pct = paste0(round(accuracy * 100, 1), "%"),
    ci_95 = paste0("[", round(ci_lower * 100, 1), "%, ", round(ci_upper * 100, 1), "%]")
  ) |>
  select(rank, condition, accuracy_pct, ci_95, n) |>
  head(10)

kable(leaderboard_predict,
      col.names = c("Rank", "Configuration", "Accuracy", "95% CI", "N"),
      caption = "Predict Mode Leaderboard (Top 10 Configurations)")
```

### Forest Plot: All Configurations

Visual comparison of all experimental factor combinations. Non-overlapping confidence intervals suggest statistically significant differences.

```{r}
#| label: predict-forest-plot
#| fig-width: 10
#| fig-height: 12

# Calculate accuracy and 95% CI for all configurations
forest_data <- predict_data |>
  mutate(
    # Create readable configuration labels
    config_label = paste0(
      model_name, " | ",
      ifelse(prompt_style == "augmented", "Aug", "Base"), " | ",
      ifelse(enable_thinking, "Think", "No-Think"), " | ",
      ifelse(vot_ray_trace, "VoT", "No-VoT")
    )
  ) |>
  group_by(config_label, model_name, prompt_style, enable_thinking, vot_ray_trace) |>
  summarise(
    n = n(),
    accuracy = mean(correct),
    # 95% CI for proportion using normal approximation
    ci_lower = accuracy - 1.96 * sqrt(accuracy * (1 - accuracy) / n),
    ci_upper = accuracy + 1.96 * sqrt(accuracy * (1 - accuracy) / n),
    .groups = "drop"
  ) |>
  # Order by accuracy (highest first)
  arrange(desc(accuracy)) |>
  mutate(
    # Create ordered factor for plotting
    config_label = factor(config_label, levels = config_label)
  )

# Forest plot
p_forest <- ggplot(forest_data, aes(x = accuracy, y = config_label)) +
  geom_point(aes(color = model_name), size = 3) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = model_name),
                 height = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50", alpha = 0.5) +
  geom_text(aes(label = paste0(round(accuracy * 100, 1), "%")),
            hjust = -0.3, size = 3) +
  labs(
    title = "Predict Mode: All Configurations Ranked by Accuracy",
    subtitle = "Error bars show 95% confidence intervals. Non-overlapping CIs suggest significant differences.",
    x = "Accuracy",
    y = "Configuration",
    color = "Model"
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    breaks = seq(0, 1, 0.1)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, family = "mono"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

print(p_forest)
```

**Total configurations**: `r nrow(forest_data)`
**Best performing**: `r as.character(forest_data$config_label[1])`
**Accuracy**: `r round(forest_data$accuracy[1] * 100, 1)`% (95% CI: [`r round(forest_data$ci_lower[1] * 100, 1)`%, `r round(forest_data$ci_upper[1] * 100, 1)`%])

### Factor-Level Comparisons

Shows accuracy by individual factors, collapsing across other dimensions (marginal effects).

```{r}
#| label: predict-factor-comparisons
#| fig-width: 10
#| fig-height: 8

# Calculate marginal effects for each factor
factor_data <- list(
  Model = predict_data |>
    group_by(model_name) |>
    summarise(
      n = n(),
      accuracy = mean(correct),
      ci_lower = accuracy - 1.96 * sqrt(accuracy * (1 - accuracy) / n),
      ci_upper = accuracy + 1.96 * sqrt(accuracy * (1 - accuracy) / n)
    ) |>
    mutate(factor = "Model", level = model_name),

  Prompt = predict_data |>
    group_by(prompt_style) |>
    summarise(
      n = n(),
      accuracy = mean(correct),
      ci_lower = accuracy - 1.96 * sqrt(accuracy * (1 - accuracy) / n),
      ci_upper = accuracy + 1.96 * sqrt(accuracy * (1 - accuracy) / n)
    ) |>
    mutate(factor = "Prompt", level = prompt_style),

  Thinking = predict_data |>
    group_by(enable_thinking) |>
    summarise(
      n = n(),
      accuracy = mean(correct),
      ci_lower = accuracy - 1.96 * sqrt(accuracy * (1 - accuracy) / n),
      ci_upper = accuracy + 1.96 * sqrt(accuracy * (1 - accuracy) / n)
    ) |>
    mutate(factor = "Extended Thinking", level = ifelse(enable_thinking, "Enabled", "Disabled")),

  VoT = predict_data |>
    group_by(vot_ray_trace) |>
    summarise(
      n = n(),
      accuracy = mean(correct),
      ci_lower = accuracy - 1.96 * sqrt(accuracy * (1 - accuracy) / n),
      ci_upper = accuracy + 1.96 * sqrt(accuracy * (1 - accuracy) / n)
    ) |>
    mutate(factor = "VoT Ray Trace", level = ifelse(vot_ray_trace, "Enabled", "Disabled"))
) |>
  bind_rows() |>
  mutate(factor = factor(factor, levels = c("Model", "Prompt", "Extended Thinking", "VoT Ray Trace")))

# Factor comparison plot
p_factors <- ggplot(factor_data, aes(x = accuracy, y = level)) +
  geom_point(size = 4, color = "#2C3E50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.2, linewidth = 1, color = "#2C3E50") +
  geom_text(aes(label = paste0(round(accuracy * 100, 1), "%")),
            hjust = -0.3, size = 3.5) +
  facet_wrap(~factor, ncol = 1, scales = "free_y") +
  labs(
    title = "Main Effects: Accuracy by Individual Factors",
    subtitle = "Marginal effects averaging across all other factors",
    x = "Accuracy",
    y = NULL
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    breaks = seq(0, 1, 0.1)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

print(p_factors)
```

```{r}
#| label: predict-by-config

# Accuracy by configuration to assess difficulty variation
predict_by_config <- predict_data |>
  group_by(config_index, model_name) |>
  summarise(
    n = n(),
    accuracy = mean(correct),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = model_name, values_from = accuracy) |>
  mutate(across(where(is.numeric) & !config_index, \(x) round(x, 3)))

kable(predict_by_config,
      caption = "Prediction Accuracy by Configuration and Model")
```

### ANOVA: Prediction Accuracy

This analysis uses mixed-effects logistic regression to test whether experimental factors significantly affect prediction accuracy, while accounting for clustering within configurations and individual rays.

**Model specification**:
- **Outcome**: Correct prediction (binary: 0/1)
- **Fixed effects**: Model, prompt style, extended thinking, VoT ray trace, and their interactions
- **Covariates**: cells_traveled (path length), atoms_affecting (atom density)
- **Random effects**: Nested structure with config_index (configuration-level) and ray_id nested within config (ray-level)

**Why nested random effects?**

- **Configuration-level (config_index)**: Accounts for overall difficulty due to specific atom placements; allows generalization beyond these 10 configurations
- **Ray-level (ray_id:config_index)**: Accounts for individual ray difficulty; handles correlation when the same ray is tested across multiple experimental conditions
- **Why nested?**: ray_id "NORTH_3" represents different physical rays across configurations (different atom placements), so must be nested, not crossed

**Covariates** (cells_traveled, atoms_affecting) are **NOT confounders** (same rays tested across all conditions) but **ARE precision variables** that explain variance and improve statistical power.

```{r}
#| label: predict-anova

# Convert factors for ANOVA
predict_data <- predict_data |>
  mutate(
    model_name = factor(model_name, levels = c("Haiku 4.5", "Sonnet 4.5", "Opus 4.5")),
    prompt_style = factor(prompt_style, levels = c("baseline", "augmented")),
    enable_thinking = factor(enable_thinking, levels = c(FALSE, TRUE)),
    vot_ray_trace = factor(vot_ray_trace, levels = c(FALSE, TRUE)),
    config_index = factor(config_index),
    # Create unique ray identifier (nested within config)
    ray_id = paste(toupper(ray_entry$side), ray_entry$pos, sep = "_")
  )

# Build model formula dynamically based on available factor levels
has_model_var <- length(unique(predict_data$model_name)) > 1
has_prompt_var <- length(unique(predict_data$prompt_style)) > 1
has_thinking_var <- length(unique(predict_data$enable_thinking)) > 1
has_vot_var <- length(unique(predict_data$vot_ray_trace)) > 1

# Build formula components
formula_parts <- c()
if (has_model_var) formula_parts <- c(formula_parts, "model_name")
if (has_prompt_var) formula_parts <- c(formula_parts, "prompt_style")
if (has_thinking_var) formula_parts <- c(formula_parts, "enable_thinking")

# Add interactions only if all three main effects are present
if (has_model_var && has_prompt_var && has_thinking_var) {
  main_formula <- "model_name * prompt_style * enable_thinking"
} else {
  main_formula <- paste(formula_parts, collapse = " + ")
}

# Add VoT if it varies
if (has_vot_var) {
  full_formula <- paste0("correct ~ ", main_formula, " + vot_ray_trace + cells_traveled + atoms_affecting + (1 | config_index/ray_id)")
} else {
  full_formula <- paste0("correct ~ ", main_formula, " + cells_traveled + atoms_affecting + (1 | config_index/ray_id)")
}

model_predict <- glmer(as.formula(full_formula),
                       family = binomial(link = "logit"),
                       data = predict_data,
                       control = glmerControl(optimizer = "bobyqa"))
```

**Model formula**: `r full_formula`

```{r}
#| label: predict-anova-results
#| echo: false

print(Anova(model_predict, type = 3))
```

#### Odds Ratios with 95% Confidence Intervals

**Interpretation**: OR > 1 means the factor increases odds of correct prediction; OR < 1 means it decreases odds. These are conditional odds ratios (controlling for random effects).

```{r}
#| label: predict-odds-ratios
#| echo: false

# Get odds ratios with confidence intervals and p-values
fixed_coefs <- fixef(model_predict)
conf_int <- confint(model_predict, parm = "beta_", method = "Wald", quiet = TRUE)

# Extract p-values using tidy() from broom.mixed
tidy_results <- tidy(model_predict, effects = "fixed") |>
  select(term, p.value)

# Create odds ratios table
or_results <- exp(cbind(OR = fixed_coefs, conf_int))
or_df <- as.data.frame(or_results) |>
  rownames_to_column("Term") |>
  filter(!grepl("Intercept", Term)) |>
  left_join(tidy_results, by = c("Term" = "term")) |>
  mutate(
    # Round numeric values
    across(c(OR, `2.5 %`, `97.5 %`), ~round(., 3)),
    p.value = round(p.value, 4),
    # Determine significance
    significant = p.value < 0.05
  )

# Format for display - add significance indicator
or_df_display <- or_df |>
  mutate(
    `Odds Ratio` = OR,
    `95% CI Lower` = `2.5 %`,
    `95% CI Upper` = `97.5 %`,
    `p-value` = format.pval(p.value, digits = 3, eps = 0.001),
    Sig = ifelse(significant, "*", "")
  ) |>
  select(Term, `Odds Ratio`, `95% CI Lower`, `95% CI Upper`, `p-value`, Sig)

# Create table
kable(or_df_display,
      caption = "Odds Ratios for Prediction Accuracy (Fixed Effects). * indicates p < 0.05.")
```

#### Random Effects Variance Components

**Configuration-level random effects** (config_index):
- Accounts for overall difficulty due to specific atom placements
- Generalizes inference beyond these 10 configurations
- Captures configuration-specific effects (e.g., dense vs sparse atom patterns)

**Ray-level random effects** (ray_id:config_index):
- Accounts for individual ray difficulty within each configuration
- Handles correlation: Same ray tested across multiple conditions
- Allows each ray to have its own baseline difficulty
- Essential when same rays appear in multiple experimental conditions

```{r}
#| label: predict-random-effects
#| echo: false

print(VarCorr(model_predict))

icc_value <- performance::icc(model_predict)
```

**Intraclass Correlation Coefficient (ICC)**: `r round(icc_value$ICC_adjusted, 3)` — Proportion of variance due to clustering at both configuration and ray-within-config levels.

```{r}
#| label: predict-model-fit
#| echo: false

r2_result <- performance::r2_nakagawa(model_predict)
```

**Model Fit** (Pseudo R², Nakagawa & Schielzeth):
- **Marginal R²** (fixed effects only): `r round(r2_result$R2_marginal, 3)`
- **Conditional R²** (fixed + random effects): `r round(r2_result$R2_conditional, 3)`

#### Cramér's V (Effect Size for Categorical Associations)

Interpretation guidelines: 0.10=small, 0.30=medium, 0.50=large

```{r}
#| label: predict-cramers-v
#| echo: false

cramers_results <- tibble(Factor = character(), `Cramér's V` = numeric())

if (has_model_var) {
  cramers_v_model <- cramers_v(table(predict_data$model_name, predict_data$correct))
  cramers_results <- cramers_results |> add_row(Factor = "Model", `Cramér's V` = cramers_v_model$Cramers_v)
}
if (has_prompt_var) {
  cramers_v_prompt <- cramers_v(table(predict_data$prompt_style, predict_data$correct))
  cramers_results <- cramers_results |> add_row(Factor = "Prompt Style", `Cramér's V` = cramers_v_prompt$Cramers_v)
}
if (has_thinking_var) {
  cramers_v_thinking <- cramers_v(table(predict_data$enable_thinking, predict_data$correct))
  cramers_results <- cramers_results |> add_row(Factor = "Extended Thinking", `Cramér's V` = cramers_v_thinking$Cramers_v)
}
if (has_vot_var) {
  cramers_v_vot <- cramers_v(table(predict_data$vot_ray_trace, predict_data$correct))
  cramers_results <- cramers_results |> add_row(Factor = "VoT Ray Trace", `Cramér's V` = cramers_v_vot$Cramers_v)
}

cramers_results <- cramers_results |> mutate(`Cramér's V` = round(`Cramér's V`, 3))

if (nrow(cramers_results) > 0) {
  kable(cramers_results, caption = "Association Strength with Prediction Accuracy")
}
```

#### Regression Coefficients for Main Effects

Key findings from mixed-effects logistic regression (fixed effects only).

```{r}
#| label: predict-regression-coefs
#| echo: false

# Get tidy output for main effects (fixed effects only)
tidy_results <- tidy(model_predict, effects = "fixed", conf.int = TRUE) |>
  filter(!grepl("Intercept|:", term)) |>  # Main effects only (no interactions)
  mutate(
    OR = exp(estimate),
    OR_lower = exp(conf.low),
    OR_upper = exp(conf.high)
  ) |>
  select(term, estimate, OR, OR_lower, OR_upper, p.value) |>
  mutate(across(c(estimate, OR, OR_lower, OR_upper), ~round(., 3)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001))

kable(tidy_results,
      col.names = c("Term", "β (log-odds)", "OR", "OR 95% CI Lower", "OR 95% CI Upper", "p-value"),
      caption = "Regression Coefficients for Main Effects (Fixed Effects)")
```

### Effect Modification by Path Complexity

Does path complexity (measured as total cells traveled: entry + internal + exit) moderate treatment effects? If treatment effects are constant across spatial complexity levels, limitations lie in inferential machinery rather than spatial processing. If effects are moderated by complexity, spatial reasoning contributes to observed failures.

```{r}
#| label: predict-effect-modification

# Only run if we have variation in factors and valid cells_traveled data
run_effect_mod <- has_model_var || has_prompt_var || has_thinking_var

if (run_effect_mod) {
  # Build interaction formula
  interaction_parts <- c()
  if (has_model_var) interaction_parts <- c(interaction_parts, "model_name * cells_traveled")
  if (has_prompt_var) interaction_parts <- c(interaction_parts, "prompt_style * cells_traveled")
  if (has_thinking_var) interaction_parts <- c(interaction_parts, "enable_thinking * cells_traveled")

  # Add nested random effects
  pred_mod_formula <- paste0("correct ~ ", paste(interaction_parts, collapse = " + "), " + (1 | config_index/ray_id)")

  model_predict_mod <- glmer(as.formula(pred_mod_formula),
                             family = binomial(link = "logit"),
                             data = predict_data,
                             control = glmerControl(optimizer = "bobyqa"))
}
```

**Model formula**: `r if(exists("pred_mod_formula")) pred_mod_formula`

```{r}
#| label: predict-effect-modification-anova
#| echo: false
#| eval: !expr run_effect_mod

# ANOVA table
print(Anova(model_predict_mod, type = 3))
```

**Interpretation of interactions**:
- **No interaction**: Treatment effect is constant across simple and complex paths
- **Positive interaction**: Treatment effect increases with path complexity
- **Negative interaction**: Treatment effect decreases with path complexity

```{r}
#| label: predict-effect-modification-coefs
#| echo: false
#| eval: !expr run_effect_mod

# Extract interaction coefficients
interaction_coefs <- tidy(model_predict_mod, conf.int = TRUE) |>
  filter(grepl(":", term)) |>  # Interaction terms only
  filter(grepl("cells_traveled", term)) |>
  mutate(
    OR = exp(estimate),
    OR_lower = exp(conf.low),
    OR_upper = exp(conf.high)
  ) |>
  select(term, estimate, OR, OR_lower, OR_upper, p.value) |>
  mutate(across(c(estimate, OR, OR_lower, OR_upper), ~round(., 3)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001))

if (nrow(interaction_coefs) > 0) {
  kable(interaction_coefs,
        col.names = c("Interaction Term", "β (log-odds)", "OR", "OR 95% CI Lower", "OR 95% CI Upper", "p-value"),
        caption = "Path Complexity Interaction Effects")
}

cells_summary <- predict_data |>
  summarise(
    min_cells = min(cells_traveled, na.rm = TRUE),
    q25_cells = quantile(cells_traveled, 0.25, na.rm = TRUE),
    median_cells = median(cells_traveled, na.rm = TRUE),
    q75_cells = quantile(cells_traveled, 0.75, na.rm = TRUE),
    max_cells = max(cells_traveled, na.rm = TRUE),
    mean_cells = mean(cells_traveled, na.rm = TRUE),
    sd_cells = sd(cells_traveled, na.rm = TRUE)
  ) |>
  mutate(across(everything(), ~round(., 2)))
```

**Path Complexity Distribution** (cells traveled):

| Statistic | Value |
|-----------|-------|
| Min | `r if(exists("cells_summary")) cells_summary$min_cells` |
| Q1 | `r if(exists("cells_summary")) cells_summary$q25_cells` |
| Median | `r if(exists("cells_summary")) cells_summary$median_cells` |
| Q3 | `r if(exists("cells_summary")) cells_summary$q75_cells` |
| Max | `r if(exists("cells_summary")) cells_summary$max_cells` |
| Mean | `r if(exists("cells_summary")) cells_summary$mean_cells` |
| SD | `r if(exists("cells_summary")) cells_summary$sd_cells` |

```{r}
#| echo: false

if (!run_effect_mod) {
  cat("Insufficient factor variation for effect modification analysis.\n")
}
```

### Effect Modification by Atom Density

Does atom density (number of atoms adjacent to ray path) moderate treatment effects?

```{r}
#| label: predict-effect-modification-atoms

run_effect_mod_atoms <- has_model_var || has_prompt_var || has_thinking_var

if (run_effect_mod_atoms) {
  # Build interaction formula
  interaction_parts <- c()
  if (has_model_var) interaction_parts <- c(interaction_parts, "model_name * atoms_affecting")
  if (has_prompt_var) interaction_parts <- c(interaction_parts, "prompt_style * atoms_affecting")
  if (has_thinking_var) interaction_parts <- c(interaction_parts, "enable_thinking * atoms_affecting")

  # Add nested random effects
  pred_mod_formula_atoms <- paste0("correct ~ ", paste(interaction_parts, collapse = " + "), " + (1 | config_index/ray_id)")

  model_predict_mod_atoms <- glmer(as.formula(pred_mod_formula_atoms),
                                   family = binomial(link = "logit"),
                                   data = predict_data,
                                   control = glmerControl(optimizer = "bobyqa"))
}
```

**Model formula**: `r if(exists("pred_mod_formula_atoms")) pred_mod_formula_atoms`

Mixed-effects logistic regression tests whether treatment effects vary with atom density.

```{r}
#| label: predict-atom-density-anova
#| echo: false
#| eval: !expr run_effect_mod_atoms

# ANOVA table
print(Anova(model_predict_mod_atoms, type = 3))
```

**Interpretation of interactions**:
- **No interaction**: Treatment effect is constant across low and high atom density
- **Positive interaction**: Treatment effect increases with more atoms affecting path
- **Negative interaction**: Treatment effect decreases with more atoms affecting path

```{r}
#| label: predict-atom-density-coefs
#| echo: false
#| eval: !expr run_effect_mod_atoms

# Extract interaction coefficients
interaction_coefs_atoms <- tidy(model_predict_mod_atoms, conf.int = TRUE) |>
  filter(grepl(":", term)) |>  # Interaction terms only
  filter(grepl("atoms_affecting", term)) |>
  mutate(
    OR = exp(estimate),
    OR_lower = exp(conf.low),
    OR_upper = exp(conf.high)
  ) |>
  select(term, estimate, OR, OR_lower, OR_upper, p.value) |>
  mutate(across(c(estimate, OR, OR_lower, OR_upper), ~round(., 3)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001))

if (nrow(interaction_coefs_atoms) > 0) {
  kable(interaction_coefs_atoms,
        col.names = c("Interaction Term", "β (log-odds)", "OR", "OR 95% CI Lower", "OR 95% CI Upper", "p-value"),
        caption = "Atom Density Interaction Effects")
}

atoms_summary <- predict_data |>
  summarise(
    min_atoms = min(atoms_affecting, na.rm = TRUE),
    q25_atoms = quantile(atoms_affecting, 0.25, na.rm = TRUE),
    median_atoms = median(atoms_affecting, na.rm = TRUE),
    q75_atoms = quantile(atoms_affecting, 0.75, na.rm = TRUE),
    max_atoms = max(atoms_affecting, na.rm = TRUE),
    mean_atoms = mean(atoms_affecting, na.rm = TRUE),
    sd_atoms = sd(atoms_affecting, na.rm = TRUE)
  ) |>
  mutate(across(everything(), ~round(., 2)))
```

**Atom Density Distribution**:

| Statistic | Value |
|-----------|-------|
| Min | `r if(exists("atoms_summary")) atoms_summary$min_atoms` |
| Q1 | `r if(exists("atoms_summary")) atoms_summary$q25_atoms` |
| Median | `r if(exists("atoms_summary")) atoms_summary$median_atoms` |
| Q3 | `r if(exists("atoms_summary")) atoms_summary$q75_atoms` |
| Max | `r if(exists("atoms_summary")) atoms_summary$max_atoms` |
| Mean | `r if(exists("atoms_summary")) atoms_summary$mean_atoms` |
| SD | `r if(exists("atoms_summary")) atoms_summary$sd_atoms` |

```{r}
#| echo: false

if (!run_effect_mod_atoms) {
  cat("Insufficient factor variation for effect modification analysis.\n")
}
```

```{r}
#| label: predict-visualizations
#| fig-width: 10
#| fig-height: 6

# Accuracy by model and condition
p1_pred <- predict_data |>
  group_by(model_name, prompt_style, enable_thinking) |>
  summarise(accuracy = mean(correct),
            se = sqrt(accuracy * (1-accuracy) / n()),
            .groups = "drop") |>
  ggplot(aes(x = model_name, y = accuracy, fill = enable_thinking)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = accuracy - se, ymax = accuracy + se),
                position = position_dodge(0.8), width = 0.2) +
  facet_wrap(~prompt_style) +
  labs(title = "Predict Mode: Accuracy by Model, Thinking, and Prompt Style",
       x = "Model", y = "Accuracy", fill = "Extended\nThinking") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p1_pred)

# Accuracy by configuration (difficulty analysis)
p2_pred <- predict_data |>
  group_by(config_index, model_name) |>
  summarise(accuracy = mean(correct), .groups = "drop") |>
  ggplot(aes(x = config_index, y = accuracy, color = model_name, group = model_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Prediction Accuracy Across Configurations",
       subtitle = "Variation indicates configuration difficulty differences",
       x = "Configuration Index", y = "Accuracy", color = "Model") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal()
print(p2_pred)

# VoT Ray Trace effect
p3_pred <- predict_data |>
  group_by(model_name, vot_ray_trace) |>
  summarise(accuracy = mean(correct),
            se = sqrt(accuracy * (1-accuracy) / n()),
            .groups = "drop") |>
  ggplot(aes(x = model_name, y = accuracy, fill = vot_ray_trace)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = accuracy - se, ymax = accuracy + se),
                position = position_dodge(0.8), width = 0.2) +
  labs(title = "Effect of VoT Ray Trace Visualization",
       x = "Model", y = "Accuracy", fill = "VoT Ray\nTrace") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p3_pred)

# Path complexity vs accuracy
p4_pred <- predict_data |>
  filter(!is.na(cells_traveled)) |>
  group_by(cells_traveled, model_name) |>
  summarise(accuracy = mean(correct),
            n = n(),
            .groups = "drop") |>
  ggplot(aes(x = cells_traveled, y = accuracy, color = model_name)) +
  geom_point(aes(size = n), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  labs(title = "Performance vs Path Complexity",
       subtitle = "Does accuracy decline as rays travel through more cells?",
       x = "Cells Traveled (Entry + Internal + Exit)",
       y = "Accuracy",
       color = "Model",
       size = "N") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal()
print(p4_pred)
```

### Ray Path Complexity Analysis

```{r}
#| label: predict-complexity-analysis
#| fig-width: 10
#| fig-height: 8

# 1. Correlation between complexity measures
complexity_cor <- cor(predict_data$cells_traveled, predict_data$atoms_affecting,
                      use = "complete.obs")
```

**Correlation** between cells_traveled and atoms_affecting: **`r round(complexity_cor, 3)`**

```{r}
#| label: predict-complexity-plots
#| echo: false

# 2. Scatterplot: cells_traveled vs atoms_affecting
p_complexity_scatter <- predict_data |>
  filter(!is.na(cells_traveled) & !is.na(atoms_affecting)) |>
  ggplot(aes(x = cells_traveled, y = atoms_affecting)) +
  geom_jitter(aes(color = correct), alpha = 0.4, width = 0.2, height = 0.2) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  labs(title = "Path Length vs Atom Density",
       subtitle = paste0("Correlation: r = ", round(complexity_cor, 3)),
       x = "Cells Traveled (Entry + Internal + Exit)",
       y = "Atoms Affecting Path",
       color = "Correct") +
  scale_color_manual(values = c("TRUE" = "#00BA38", "FALSE" = "#F8766D")) +
  theme_minimal()
print(p_complexity_scatter)

# 3. Accuracy by atom density categories
p_atoms_accuracy <- predict_data |>
  filter(!is.na(atoms_affecting)) |>
  mutate(atom_category = factor(atoms_affecting,
                                 levels = 0:4,
                                 labels = c("0 atoms", "1 atom", "2 atoms", "3 atoms", "4 atoms"))) |>
  group_by(atom_category, model_name) |>
  summarise(
    accuracy = mean(correct),
    se = sqrt(accuracy * (1-accuracy) / n()),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = atom_category, y = accuracy, fill = model_name)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = accuracy - se, ymax = accuracy + se),
                position = position_dodge(0.8), width = 0.2) +
  geom_text(aes(label = paste0("n=", n)),
            position = position_dodge(0.8), vjust = -0.5, size = 3) +
  labs(title = "Accuracy by Atom Density",
       subtitle = "How many atoms are adjacent to the ray path?",
       x = "Number of Atoms Affecting Ray Path",
       y = "Accuracy",
       fill = "Model") +
  scale_y_continuous(limits = c(0, 1.1), labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
print(p_atoms_accuracy)

# 4. Accuracy by path length quartiles
p_cells_quartiles <- predict_data |>
  filter(!is.na(cells_traveled)) |>
  mutate(
    path_quartile = cut(cells_traveled,
                        breaks = quantile(cells_traveled, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE),
                        labels = c("Q1 (Shortest)", "Q2", "Q3", "Q4 (Longest)"),
                        include.lowest = TRUE)
  ) |>
  group_by(path_quartile, model_name) |>
  summarise(
    accuracy = mean(correct),
    se = sqrt(accuracy * (1-accuracy) / n()),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = path_quartile, y = accuracy, fill = model_name)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = accuracy - se, ymax = accuracy + se),
                position = position_dodge(0.8), width = 0.2) +
  geom_text(aes(label = paste0("n=", n)),
            position = position_dodge(0.8), vjust = -0.5, size = 3) +
  labs(title = "Accuracy by Path Length Quartiles",
       subtitle = "Does accuracy decline for longer ray paths?",
       x = "Path Length Quartile",
       y = "Accuracy",
       fill = "Model") +
  scale_y_continuous(limits = c(0, 1.1), labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
print(p_cells_quartiles)
```

```{r}
#| label: predict-error-analysis

# Analyze error patterns by comparing predicted vs actual
# This helps identify systematic errors (e.g., absorption vs deflection confusion)

# First, extract the type of result from predicted and actual
predict_errors <- predict_data |>
  filter(!correct) |>
  mutate(
    predicted_type = case_when(
      grepl("absorbed|HIT", predicted, ignore.case = TRUE) ~ "absorbed",
      grepl("reflected|REFLECTION", predicted, ignore.case = TRUE) ~ "reflected",
      TRUE ~ "detour"
    ),
    actual_type = case_when(
      grepl("absorbed|HIT", actual, ignore.case = TRUE) ~ "absorbed",
      grepl("reflected|REFLECTION", actual, ignore.case = TRUE) ~ "reflected",
      TRUE ~ "detour"
    )
  )

# Error confusion matrix
error_matrix <- predict_errors |>
  count(actual_type, predicted_type) |>
  pivot_wider(names_from = predicted_type, values_from = n, values_fill = 0)

kable(error_matrix, caption = "Prediction Errors: Actual vs Predicted Outcome Type (Rows = Actual, Columns = Predicted)")

# Error rates by model
error_by_model <- predict_errors |>
  count(model_name, actual_type, predicted_type) |>
  group_by(model_name) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

error_by_model |>
  group_by(model_name) |>
  slice_max(n, n = 3) |>
  mutate(error_pattern = paste(actual_type, "->", predicted_type)) |>
  select(model_name, error_pattern, n, prop) |>
  mutate(prop = round(prop, 3)) |>
  kable(col.names = c("Model", "Error Pattern (Actual → Predicted)", "Count", "Proportion"),
        caption = "Most Common Error Types by Model (Top 3 per Model)")
```

## Play Mode Results

```{r}
#| label: load-play-data

# Load all Play mode experiment JSON files from Experiment 1/Play/
play_json_files <- list.files(
  path = "Experiment 1/Play",
  pattern = "\\.json$",
  full.names = TRUE
)

# Function to extract results from a single JSON file
extract_play_results <- function(json_file) {
  data <- jsonlite::fromJSON(json_file, simplifyDataFrame = FALSE)

  # Extract results array
  results <- data$results

  # Convert each result to a tibble row
  map_dfr(results, function(r) {
    # Calculate mean complexity measures for each ray fired
    ray_complexity <- lapply(r$raySequence, function(ray) {
      if (!is.null(ray$action) && ray$action == "fire" &&
          !is.null(ray$rayResult) && !is.null(ray$rayResult$path)) {
        list(
          atoms = count_atoms_affecting(ray$rayResult$path, r$atomConfig),
          cells = count_ray_cells(ray$rayResult)
        )
      } else {
        list(atoms = NA_real_, cells = NA_real_)
      }
    })

    mean_atoms <- mean(sapply(ray_complexity, function(x) x$atoms), na.rm = TRUE)
    mean_cells <- mean(sapply(ray_complexity, function(x) x$cells), na.rm = TRUE)
    if (is.nan(mean_atoms)) mean_atoms <- NA_real_
    if (is.nan(mean_cells)) mean_cells <- NA_real_

    tibble(
      experiment_id = r$experimentId,
      start_time = r$startTime,
      end_time = r$endTime,
      model = r$model,
      model_name = r$modelName,
      prompt_style = r$promptStyle,
      prompt_condition = r$promptCondition,
      mode = r$mode,
      config_index = r$configIndex,
      atom_config = list(r$atomConfig),
      include_visualization = r$includeVisualization,
      allow_hypotheses = r$allowHypotheses,
      enable_thinking = r$enableThinking,
      thinking_budget = r$thinkingBudget,
      vot_grid_state = r$votGridState,
      vot_ray_trace = r$votRayTrace,
      vot_hypothesis = r$votHypothesis,
      rays_used = r$raysUsed,
      atoms_correct = r$atomsCorrect,
      atoms_missed = r$atomsMissed,
      score = r$score,
      invalid_moves = r$invalidMoves,
      hypothesis_actions = r$hypothesisActions,
      total_api_calls = r$totalApiCalls,
      total_input_tokens = r$totalInputTokens,
      total_output_tokens = r$totalOutputTokens,
      total_response_time_ms = r$totalResponseTimeMs,
      mean_atoms_per_ray = mean_atoms,
      mean_cells_per_ray = mean_cells,
      source_file = basename(json_file)
    )
  })
}

# Load all results into a single long-format tibble
play_data <- map_dfr(play_json_files, extract_play_results)
```

Data loaded: `r nrow(play_data)` experiment results from `r length(play_json_files)` JSON files.

- **Models**: `r paste(unique(play_data$model_name), collapse = ", ")`
- **Prompt styles**: `r paste(unique(play_data$prompt_style), collapse = ", ")`
- **Configurations tested**: `r length(unique(play_data$config_index))`
- **Mean atoms per ray range**: `r round(min(play_data$mean_atoms_per_ray, na.rm = TRUE), 2)`–`r round(max(play_data$mean_atoms_per_ray, na.rm = TRUE), 2)`
- **Mean cells per ray range**: `r round(min(play_data$mean_cells_per_ray, na.rm = TRUE), 2)`–`r round(max(play_data$mean_cells_per_ray, na.rm = TRUE), 2)`

```

```{r}
#| label: play-results-summary

# Summary statistics by model, prompt, and thinking
play_summary <- play_data |>
  group_by(model_name, prompt_style, enable_thinking) |>
  summarise(
    n = n(),
    atoms_correct_mean = mean(atoms_correct),
    atoms_correct_sd = sd(atoms_correct),
    score_mean = mean(score),
    score_sd = sd(score),
    mean_atoms_ray = mean(mean_atoms_per_ray, na.rm = TRUE),
    rays_used_mean = mean(rays_used),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !n, \(x) round(x, 2)))

kable(play_summary,
      col.names = c("Model", "Prompt", "Thinking", "N",
                    "Atoms Correct (M)", "Atoms Correct (SD)",
                    "Score (M)", "Score (SD)",
                    "Atoms/Ray", "Rays Used"))
```

```{r}
#| label: play-summary-by-vot

# Summary by VoT factors
play_summary_vot <- play_data |>
  group_by(vot_grid_state, vot_ray_trace, vot_hypothesis) |>
  summarise(
    n = n(),
    atoms_correct_mean = mean(atoms_correct),
    atoms_correct_sd = sd(atoms_correct),
    score_mean = mean(score),
    score_sd = sd(score),
    mean_atoms_ray = mean(mean_atoms_per_ray, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !n, \(x) round(x, 2)))

kable(play_summary_vot,
      col.names = c("VoT Grid", "VoT Ray", "VoT Hyp", "N",
                    "Atoms Correct (M)", "Atoms Correct (SD)",
                    "Score (M)", "Score (SD)", "Atoms/Ray"))
```

### Play Mode Leaderboard

Top 10 configurations by atoms correct. Format matches typical LLM benchmark reporting for cross-study comparison.

```{r}
#| label: play-leaderboard

leaderboard_play <- play_data |>
  mutate(
    condition = paste0(
      model_name, " | ",
      ifelse(prompt_style == "augmented", "Aug", "Base"), " | ",
      ifelse(enable_thinking, "Think", "No-Think")
    )
  ) |>
  group_by(condition, model_name, prompt_style, enable_thinking) |>
  summarise(
    n = n(),
    atoms_correct_mean = mean(atoms_correct),
    atoms_correct_se = sd(atoms_correct) / sqrt(n),
    score_mean = mean(score),
    perfect_games = sum(atoms_correct == 4),
    .groups = "drop"
  ) |>
  arrange(desc(atoms_correct_mean), score_mean) |>
  mutate(
    rank = row_number(),
    atoms_ci = paste0(round(atoms_correct_mean, 2), " ± ", round(atoms_correct_se, 2)),
    perfect_pct = paste0(round(perfect_games / n * 100, 1), "%")
  ) |>
  select(rank, condition, atoms_ci, score_mean, perfect_pct, n) |>
  head(10)

kable(leaderboard_play,
      col.names = c("Rank", "Configuration", "Atoms Correct (M ± SE)", "Score (M)", "Perfect Games", "N"),
      caption = "Play Mode Leaderboard (Top 10 Configurations)")
```

### ANOVA: Play Mode Performance

Full factorial ANOVA using mixed-effects models with configuration-level random effects. Covariates (mean_atoms_per_ray, mean_cells_per_ray) capture ray difficulty.

```{r}
#| label: play-anova-analysis

# Convert factors
play_data <- play_data |>
  mutate(
    model_name = factor(model_name, levels = c("Haiku 4.5", "Sonnet 4.5", "Opus 4.5")),
    prompt_style = factor(prompt_style, levels = c("baseline", "augmented")),
    enable_thinking = factor(enable_thinking, levels = c(FALSE, TRUE)),
    vot_grid_state = factor(vot_grid_state, levels = c(FALSE, TRUE)),
    vot_ray_trace = factor(vot_ray_trace, levels = c(FALSE, TRUE)),
    vot_hypothesis = factor(vot_hypothesis, levels = c(FALSE, TRUE))
  )

# Build model formula dynamically
has_model_var <- length(unique(play_data$model_name)) > 1
has_prompt_var <- length(unique(play_data$prompt_style)) > 1
has_thinking_var <- length(unique(play_data$enable_thinking)) > 1
has_vot_grid <- length(unique(play_data$vot_grid_state)) > 1
has_vot_ray <- length(unique(play_data$vot_ray_trace)) > 1
has_vot_hyp <- length(unique(play_data$vot_hypothesis)) > 1

# Build main formula
formula_parts <- c()
if (has_model_var) formula_parts <- c(formula_parts, "model_name")
if (has_prompt_var) formula_parts <- c(formula_parts, "prompt_style")
if (has_thinking_var) formula_parts <- c(formula_parts, "enable_thinking")

if (has_model_var && has_prompt_var && has_thinking_var) {
  main_formula <- "model_name * prompt_style * enable_thinking"
} else {
  main_formula <- paste(formula_parts, collapse = " + ")
}

# Add VoT factors
vot_parts <- c()
if (has_vot_grid) vot_parts <- c(vot_parts, "vot_grid_state")
if (has_vot_ray) vot_parts <- c(vot_parts, "vot_ray_trace")
if (has_vot_hyp) vot_parts <- c(vot_parts, "vot_hypothesis")

if (length(vot_parts) > 0) {
  atoms_formula <- paste0("atoms_correct ~ ", main_formula, " + ",
                          paste(vot_parts, collapse = " + "),
                          " + mean_cells_per_ray + mean_atoms_per_ray + (1 | config_index)")
} else {
  atoms_formula <- paste0("atoms_correct ~ ", main_formula,
                          " + mean_cells_per_ray + mean_atoms_per_ray + (1 | config_index)")
}

model_atoms <- lmer(as.formula(atoms_formula), data = play_data)
```

#### Model 1: Atoms Correct (0-4)

Testing which factors affect the number of atoms correctly identified.

**Model formula**: `r atoms_formula`

```{r}
#| label: play-atoms-anova
#| echo: false

print(Anova(model_atoms, type = 3))

# Random effects variance
print(VarCorr(model_atoms))

icc_value <- performance::icc(model_atoms)
r2_result <- performance::r2_nakagawa(model_atoms)
```

**Random Effects** — Configuration (config_index) variance accounts for config-level difficulty and allows generalization beyond these 10 configurations.

**ICC**: `r round(icc_value$ICC_adjusted, 3)` — Proportion of variance due to configuration clustering

**Model Fit**:
- **Marginal R²** (fixed effects only): `r round(r2_result$R2_marginal, 3)`
- **Conditional R²** (fixed + random effects): `r round(r2_result$R2_conditional, 3)`

```{r}
#| label: play-score-model

# Model 2: Score (main effects with config random effects)
if (length(vot_parts) > 0) {
  score_formula <- paste0("score ~ ", main_formula, " + ",
                          paste(vot_parts, collapse = " + "),
                          " + mean_cells_per_ray + mean_atoms_per_ray + (1 | config_index)")
} else {
  score_formula <- paste0("score ~ ", main_formula,
                          " + mean_cells_per_ray + mean_atoms_per_ray + (1 | config_index)")
}

model_score <- lmer(as.formula(score_formula), data = play_data)
```

#### Model 2: Score (Lower is Better)

Testing which factors affect game score.

**Model formula**: `r score_formula`

```{r}
#| label: play-score-anova
#| echo: false

print(Anova(model_score, type = 3))

# Random effects
print(VarCorr(model_score))

icc_score <- performance::icc(model_score)
r2_score <- performance::r2_nakagawa(model_score)
```

**ICC**: `r round(icc_score$ICC_adjusted, 3)`

**Model Fit**:
- **Marginal R²**: `r round(r2_score$R2_marginal, 3)`
- **Conditional R²**: `r round(r2_score$R2_conditional, 3)`

```

### Effect Modification by Ray Complexity (Play Mode)

Does ray complexity (mean atoms per ray) moderate treatment effects?

```{r}
#| label: play-effect-modification

run_play_effect_mod <- has_model_var || has_prompt_var || has_thinking_var

if (run_play_effect_mod) {
  # Build interaction formula with random effect
  interaction_parts <- c()
  if (has_model_var) interaction_parts <- c(interaction_parts, "model_name * mean_atoms_per_ray")
  if (has_prompt_var) interaction_parts <- c(interaction_parts, "prompt_style * mean_atoms_per_ray")
  if (has_thinking_var) interaction_parts <- c(interaction_parts, "enable_thinking * mean_atoms_per_ray")

  atoms_mod_formula <- paste0("atoms_correct ~ ",
                              paste(interaction_parts, collapse = " + "),
                              " + (1 | config_index)")
  score_mod_formula <- paste0("score ~ ",
                              paste(interaction_parts, collapse = " + "),
                              " + (1 | config_index)")

  model_atoms_mod <- lmer(as.formula(atoms_mod_formula), data = play_data)
  model_score_mod <- lmer(as.formula(score_mod_formula), data = play_data)
}
```

**Atoms Correct Model**: `r if(exists("atoms_mod_formula")) atoms_mod_formula`

```{r}
#| label: play-effect-mod-atoms
#| echo: false
#| eval: !expr run_play_effect_mod

print(Anova(model_atoms_mod, type = 3))
```

**Score Model**: `r if(exists("score_mod_formula")) score_mod_formula`

```{r}
#| label: play-effect-mod-score
#| echo: false
#| eval: !expr run_play_effect_mod

print(Anova(model_score_mod, type = 3))
```

```{r}
#| echo: false

if (!run_play_effect_mod) {
  cat("Insufficient factor variation for effect modification analysis.\n")
}
```

### Regression Analysis: Play Mode

Regression complements ANOVA by providing interpretable effect magnitudes. Coefficients show the expected change in outcome per unit change in predictor.

```{r}
#| label: play-regression-analysis

# Build formula with factors that have variation
regression_parts <- formula_parts  # From earlier: model_name, prompt_style, enable_thinking
if (has_vot_grid) regression_parts <- c(regression_parts, "vot_grid_state")
if (has_vot_ray) regression_parts <- c(regression_parts, "vot_ray_trace")
if (has_vot_hyp) regression_parts <- c(regression_parts, "vot_hypothesis")
regression_parts <- c(regression_parts, "mean_atoms_per_ray", "mean_cells_per_ray")

atoms_reg_formula <- paste0("atoms_correct ~ ", paste(regression_parts, collapse = " + "),
                            " + (1 | config_index)")

lmer_atoms <- lmer(as.formula(atoms_reg_formula), data = play_data)

# Tidy summary (fixed effects only)
atoms_reg_summary <- tidy(lmer_atoms, effects = "fixed", conf.int = TRUE) |>
  mutate(across(c(estimate, conf.low, conf.high, statistic), ~round(., 3)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high)

kable(atoms_reg_summary,
      col.names = c("Term", "β (Coefficient)", "SE", "t-statistic", "p-value", "95% CI Lower", "95% CI Upper"),
      caption = "Regression Coefficients: Atoms Correct (Fixed Effects)")

r2_atoms_reg <- performance::r2_nakagawa(lmer_atoms)

# Model for Score
score_reg_formula <- paste0("score ~ ", paste(regression_parts, collapse = " + "),
                            " + (1 | config_index)")

lmer_score <- lmer(as.formula(score_reg_formula), data = play_data)

score_reg_summary <- tidy(lmer_score, effects = "fixed", conf.int = TRUE) |>
  mutate(across(c(estimate, conf.low, conf.high, statistic), ~round(., 3)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) |>
  select(term, estimate, std.error, statistic, p.value, conf.low, conf.high)

kable(score_reg_summary,
      col.names = c("Term", "β (Coefficient)", "SE", "t-statistic", "p-value", "95% CI Lower", "95% CI Upper"),
      caption = "Regression Coefficients: Score (Fixed Effects)")

r2_score_reg <- performance::r2_nakagawa(lmer_score)

# Standardized coefficients for comparison
std_coef_atoms <- standardize_parameters(lmer_atoms, method = "refit")
std_coef_score <- standardize_parameters(lmer_score, method = "refit")
```

**Model Fit — Atoms Correct**:
- Marginal R²: `r round(r2_atoms_reg$R2_marginal, 3)`
- Conditional R²: `r round(r2_atoms_reg$R2_conditional, 3)`

**Model Fit — Score**:
- Marginal R²: `r round(r2_score_reg$R2_marginal, 3)`
- Conditional R²: `r round(r2_score_reg$R2_conditional, 3)`

#### Standardized Coefficients (Beta Weights)

Allows comparison of relative importance across fixed effects (based on fixed effects only).

**Atoms Correct (Top 10)**:

```{r}
#| label: play-std-coefs-atoms
#| echo: false

print(std_coef_atoms |>
      filter(!grepl("Intercept", Parameter)) |>
      arrange(desc(abs(Std_Coefficient))) |>
      head(10))
```

**Score (Top 10)**:

```{r}
#| label: play-std-coefs-score
#| echo: false

print(std_coef_score |>
      filter(!grepl("Intercept", Parameter)) |>
      arrange(desc(abs(Std_Coefficient))) |>
      head(10))
```

### Interpretation: Why Configuration Random Effects?

**Play mode data structure**:
- Unit of analysis: Entire games (one game per config × condition)
- LLM chooses which rays to fire (endogenous ray selection)
- Different conditions may fire different rays
- Multiple games per configuration (clustering)

**Why config-level random effects?**

1. Games within same config share atom placement
2. Some configs are inherently harder/easier (random variation)
3. Generalizes inference beyond these 10 specific configurations
4. Accounts for correlation among games within same config

**Why NOT ray-level random effects (unlike Predict mode)?**

1. Rays are not predetermined—LLM chooses them
2. Different conditions may fire different rays
3. Ray selection is part of the strategy (endogenous)
4. No repeated measures at individual ray level

**Covariates** (mean_cells_per_ray, mean_atoms_per_ray):
- **NOT confounders**: Ray choice is endogenous to experimental condition
- **ARE precision variables**: Describe difficulty of rays LLM chose to fire
- **Improve model fit**: Explain variance, increase power

**Impact of ignoring config-level clustering**:
- Underestimates standard errors
- Inflates Type I error (false positives)
- Invalid inference (cannot generalize beyond these 10 configs)

```

```{r}
#| label: play-visualizations
#| fig-width: 10
#| fig-height: 6

# Atoms correct by model and thinking
p1 <- ggplot(play_data, aes(x = model_name, y = atoms_correct, fill = enable_thinking)) +
  geom_boxplot() +
  facet_wrap(~prompt_style) +
  labs(title = "Atoms Correct by Model, Thinking, and Prompt Style",
       x = "Model", y = "Atoms Correct (0-4)", fill = "Thinking") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p1)

# Score distribution
p2 <- ggplot(play_data, aes(x = model_name, y = score, fill = enable_thinking)) +
  geom_boxplot() +
  facet_wrap(~prompt_style) +
  labs(title = "Score by Model, Thinking, and Prompt Style",
       x = "Model", y = "Score (lower is better)", fill = "Thinking") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p2)

# Mean atoms per ray relationship with performance
p3 <- ggplot(play_data, aes(x = mean_atoms_per_ray, y = atoms_correct, color = model_name)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Performance vs Ray Complexity",
       x = "Mean Atoms Affecting Each Ray", y = "Atoms Correct",
       color = "Model") +
  theme_minimal()
print(p3)

# Interaction plot: Prompt Style × Thinking
p4 <- ggplot(play_data, aes(x = prompt_style, y = atoms_correct,
                            color = enable_thinking, group = enable_thinking)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1) +
  labs(title = "Prompt Style × Thinking Interaction",
       x = "Prompt Style", y = "Atoms Correct (mean ± SE)",
       color = "Thinking") +
  theme_minimal()
print(p4)
```

## Error Analysis

```{r}
#| label: error-analysis-placeholder
#| eval: false

# Placeholder for error taxonomy
# Will categorize errors from extended thinking traces

# error_categories <- c(
#   "Constraint tracking failure",
#   "Spatial reasoning error",
#   "Non-parsimonious hypothesis",
#   "Failure to update on evidence",
#   "Experiment design inefficiency"
# )
```

\[Error analysis to be added after data collection\]

# Discussion

## Summary of Findings

\[To be completed after data collection\]

## Relation to Literature

### Constraint Satisfaction

\[Discuss how results relate to ZebraLogic, GenCP findings\]

### Spatial Reasoning

\[Discuss how results relate to PLUGH, spatial reasoning benchmarks\]

### World Models

\[Discuss how results relate to pulley system, text world simulator
findings\]

### Abductive Reasoning

\[Discuss how results relate to Occam's Razor, hypothesis generation
findings\]

## Implications

### For LLM Deployment in Diagnostic Domains

\[Discuss implications for medical diagnosis, fault detection, and other
diagnostic tasks\]

### For LLM-Modulo Architectures

\[Discuss whether tool augmentation addresses the limitations\]

### For Benchmark Development

\[Discuss Black Box as a diagnostic tool for reasoning assessment\]

## Limitations

1.  **Single game domain**: Results may not generalize to all reasoning
    tasks
2.  **Prompt sensitivity**: Different prompt formulations might yield
    different results
3.  **Model versions**: Rapid model updates may change performance
    characteristics
4.  **Sample size**: \[Discuss statistical power\]

## Future Directions

1.  **Extended model comparison**: GPT-4, Gemini, open-source models
2.  **Tool augmentation study**: Provide ray-tracing verification tool
3.  **Difficulty manipulation**: Vary grid size, atom count, atom
    arrangement
4.  **Human baseline**: Compare LLM performance to human players
5.  **Fine-tuning study**: Following the approach of
    @wang2025mastermind, who demonstrated substantial reasoning
    improvements through game-specific training on Doudizhu and Go, test
    whether task-specific training on Black Box traces improves
    performance. Of particular interest is whether such training would
    generalize to novel atom configurations (in-distribution transfer)
    and to other spatial reasoning tasks (out-of-distribution transfer),
    or whether improvements would be limited to memorized patterns—the
    "catastrophic forgetting" phenomenon observed in their work when
    game training interfered with unrelated reasoning tasks

# Conclusion

\[To be completed after data collection\]

# References

# Appendix A: Prompt Texts {.appendix}

## Baseline Play Prompt

```         
You are playing Black Box, a game of hide and seek played on an 8 by 8 grid.

Your opponent has hidden 4 balls within this box. By shooting rays into the
box and observing where they emerge, it is possible to deduce the positions
of the hidden balls.

GRID: 8x8, rows 1-8 (top to bottom), columns 1-8 (left to right).
RAYS: Fire from edge positions - NORTH/SOUTH use columns 1-8, EAST/WEST use rows 1-8.

There are three possible outcomes for each ray you send into the box:

DETOUR: The ray is deflected and emerges somewhere other than where you sent
it in. Detours are denoted by matching pairs of numbers.

REFLECTION (R): The ray is reflected and emerges in the same place it was sent in.

HIT (H): The ray strikes a ball directly and is absorbed.

[ASCII diagram examples showing deflection, reflection, and hit patterns...]

SCORING:
Your goal is to minimize your score. Lower is better.
- Each ray entry point costs 1 point
- Each ray exit point costs 1 point (detours cost 2 total, reflections cost 1)
- Each missed atom costs 5 points

RULES:
- You cannot fire from positions already used as entry or exit points.
- Maximum 20 rays.

Respond with JSON only:
{"action": "fire", "side": "north|south|east|west", "position": 1-8, "reasoning": "..."}
{"action": "guess", "atoms": [[row,col], [row,col], [row,col], [row,col]], "reasoning": "..."}
```

## Baseline Predict Prompt

```         
Predict where a ray will exit in Black Box.

GRID: 8x8, rows 1-8 (top to bottom), columns 1-8 (left to right).
EDGES: NORTH/SOUTH use columns 1-8, EAST/WEST use rows 1-8.

As a ray approaches a ball it is deflected ninety degrees. Rays can be
deflected multiple times.

[ASCII diagram examples showing deflection, reflection, and hit patterns...]

Respond with JSON only:
{"exit_side": "north|south|east|west", "exit_position": 1-8, "reasoning": "..."}
OR for absorption: {"absorbed": true, "reasoning": "..."}
OR for reflection: {"reflected": true, "reasoning": "..."}
```

## Augmented Prompt

\[Full augmented prompt text to be included\]

## Hypothesis-Enabled Play Prompt Extension

When hypotheses are enabled, the response format is extended:

```         
Respond with JSON only:
{"action": "fire", "side": "north|south|east|west", "position": 1-8, "reasoning": "..."}
{"action": "mark", "row": 1-8, "col": 1-8, "reasoning": "..."} - mark suspected atom
{"action": "unmark", "row": 1-8, "col": 1-8, "reasoning": "..."} - remove marked position
{"action": "check", "reasoning": "..."} - submit answer (requires exactly 4 marked positions)
```

# Appendix B: Sample Game Traces {.appendix}

\[Representative game traces showing LLM reasoning\]

# Appendix C: Error Taxonomy {.appendix}

\[Detailed coding scheme for error analysis\]
