---
title: "Black Box as a Diagnostic Benchmark for LLM Reasoning Limitations"
subtitle: "A Baseline Study of Constraint Satisfaction, Spatial Reasoning, and Abductive Inference"
author:
  - name: "Todd R. Johnson"
    affiliation: "University of Texas Health Science Center at Houston"
    email: "todd.r.johnson@uth.tmc.edu"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
bibliography: references.bib
execute:
  echo: true
  warning: false
abstract: |
  Large Language Models (LLMs) have demonstrated impressive capabilities across many domains, yet fundamental questions remain about their ability to perform systematic reasoning on novel problems. This study uses the classic Black Box game as a diagnostic benchmark to assess LLM capabilities across multiple cognitive dimensions: constraint satisfaction, spatial reasoning, world model simulation, abductive inference, and hypothesis updating. Black Box is particularly well-suited for this purpose because it requires integrating observations across time, maintaining belief states about hidden variables, and designing informative experiments—capabilities that are central to scientific and diagnostic reasoning. We present baseline results comparing Claude model variants (Haiku, Sonnet, Opus) across prompt conditions ranging from minimal human-equivalent instructions to augmented strategic guidance. Our findings reveal systematic limitations that persist despite scaling and prompt engineering, with implications for the deployment of LLMs in domains requiring novel problem-solving.
keywords: [large language models, reasoning, constraint satisfaction, spatial reasoning, abductive inference, benchmark]
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)

# Set plotting defaults
theme_set(theme_minimal())
```

# Introduction

The rapid advancement of Large Language Models (LLMs) has sparked both
enthusiasm and concern about their reasoning capabilities. While these
models demonstrate impressive performance on many benchmarks, there is
growing evidence that their abilities may be more brittle than initially
assumed, particularly when confronted with problems that require
systematic, multi-step reasoning on novel configurations
[@marcus2020next; @willison2024llms].

A critical question for deploying LLMs in consequential domains—medical
diagnosis, scientific discovery, legal reasoning—is whether they can
genuinely reason about novel problems or whether their apparent
competence reflects sophisticated pattern matching on problems similar
to their training data. This distinction has profound implications:
diagnostic and scientific problems are, by definition, novel
configurations that require genuine inference rather than retrieval.

## The Black Box Game as a Diagnostic Benchmark

Black Box is a classic deduction game designed by Eric Solomon in 1978
[@blackbox1978]. The game presents players with an 8×8 grid containing
four hidden "atoms" at unknown locations. Players fire rays into the
grid from edge positions and observe how the rays behave: they may pass
through, be absorbed by hitting an atom directly, be deflected by
passing adjacent to an atom, or be reflected by encountering atoms on
both sides of their path.

The game is particularly well-suited as a diagnostic benchmark for LLM
reasoning because it requires the integration of multiple cognitive
capabilities:

1.  **Constraint Satisfaction**: Each ray observation constrains the
    possible locations of atoms. Players must track which configurations
    remain consistent with all observations.

2.  **Spatial Reasoning**: Understanding ray behavior requires mental
    simulation of paths through 2D space, including deflection geometry.

3.  **World Model Simulation**: Predicting ray outcomes (the "forward"
    problem) requires accurate simulation of physics-like rules.

4.  **Abductive Inference**: Determining atom locations from ray
    observations (the "inverse" problem) requires inferring hidden
    causes from observable effects.

5.  **Hypothesis Updating**: As new observations arrive, beliefs about
    atom locations must be systematically revised.

6.  **Experiment Design**: Choosing which rays to fire involves
    selecting observations that will maximally disambiguate among
    competing hypotheses.

Crucially, while the rules of Black Box are simple and can be fully
specified in a prompt, any particular game configuration is almost
certainly not in the model's training data. This makes Black Box an
ideal out-of-distribution test of genuine reasoning capabilities.

## Research Questions

This study addresses the following questions:

1.  **Forward Reasoning**: Can LLMs accurately simulate ray behavior
    when given atom positions (world model accuracy)?

2.  **Inverse Reasoning**: Can LLMs correctly infer atom positions from
    ray observations (abductive inference)?

3.  **Prompt Effects**: Do augmented instructions and visualizations
    improve performance over baseline human-equivalent prompts?

4.  **Scaling Effects**: Do larger models (Opus \> Sonnet \> Haiku) show
    improved performance, and if so, is improvement sufficient to reach
    human-level competence?

5.  **Error Patterns**: What types of reasoning failures occur, and what
    do they reveal about underlying limitations?

# Literature Review

## Constraint Satisfaction and LLMs

Constraint satisfaction problems (CSPs) require finding values for
variables that satisfy a set of constraints. Classic examples include
Sudoku, scheduling problems, and logic puzzles. Recent research has
systematically evaluated LLM performance on such tasks.

The ZebraLogic benchmark [@lin2024zebralogic] evaluates LLMs on logic
grid puzzles (also known as Zebra puzzles), a well-established CSP
format used in standardized tests like the LSAT. Results indicate that
while LLMs show some capability on simple puzzles, performance degrades
substantially as problem complexity increases. The benchmark reveals
that LLMs lack several abilities required for complex logical reasoning,
including counterfactual thinking, reflective reasoning, and
compositional generalization.

@bonlarron2024gencp demonstrate that LLMs "excel at generating fluent
text but struggle to enforce external constraints because they generate
tokens sequentially without explicit control mechanisms." Their GenCP
framework addresses this by combining LLM predictions with constraint
programming (CP) reasoning, treating text generation as a CSP. This
hybrid approach significantly outperforms pure LLM methods, suggesting
that constraint satisfaction may require symbolic reasoning components
that LLMs lack.

@michailidis2024constraint evaluated LLMs on translating natural
language problem descriptions into formal constraint models. They found
that zero-shot prompting yields essentially no success, and even
few-shot prompting produces inconsistent results. This aligns with
@freuder2024conversational's observation that successful constraint
satisfaction often requires iterative refinement through dialogue, not
single-shot generation.

The implications for Black Box are direct: tracking which cell locations
are ruled out by each ray observation is fundamentally a constraint
satisfaction problem. If LLMs struggle with CSPs generally, they should
struggle with the constraint-tracking aspect of Black Box.

## Spatial Reasoning Limitations

Spatial reasoning—the ability to mentally represent and manipulate
spatial relationships—is a well-documented weakness of current LLMs.

The PLUGH benchmark [@tikhonov2024plugh] evaluates spatial understanding
and reasoning using text adventure game scenarios that require building
mental maps from textual descriptions. Results show that while some
commercial LLMs exhibit moderate reasoning abilities, "all models still
have significant room for improvement." The benchmark identifies typical
failure modes including difficulty maintaining consistent spatial
representations across multiple textual updates.

@chen2024spatial conducted an empirical analysis of spatial reasoning in
large multimodal models, finding that even GPT-4o achieves only 27.5%
accuracy on questions requiring reasoning from a human's viewpoint in an
image. Performance degrades further on multi-hop spatial reasoning
questions that require integrating multiple spatial relationships.

@bos2024spatial provides a practitioner's perspective on LLM spatial
reasoning, noting that "all of the LLMs failed immediately on the
easiest problems" involving mental box folding—a standard spatial
reasoning task in human cognitive assessments. Interestingly, LLMs
performed better on reasoning about objects in photographs than on
abstract spatial diagrams, suggesting reliance on statistical patterns
from image-caption training data rather than genuine spatial reasoning.

A synthesis of spatial reasoning research [@quan2025stark] notes that
"empirical assessments consistently show a characteristic pattern:
models exhibit moderate competence in simple, direct spatial tasks but
rapidly deteriorate as the problem scale or compositional/geometric
complexity increases, with performance losses ranging from 42% to over
80% as task complexity grows."

For Black Box, ray tracing requires understanding how rays move through
2D space and how they deflect upon encountering atoms. The documented
limitations in spatial reasoning suggest LLMs will struggle with
accurate ray path simulation.

## Reasoning Complexity and Performance Collapse

A critical question for evaluating LLM reasoning is how performance
scales with problem complexity. @shojaee2025illusion provide a
systematic investigation using controllable puzzle environments that
allow precise manipulation of compositional complexity while maintaining
consistent logical structures.

Their findings reveal three distinct performance regimes. In
low-complexity tasks, standard models surprisingly outperform reasoning
models (those using extended thinking or chain-of-thought), suggesting
that extended reasoning can introduce unnecessary complexity for simple
problems. In medium-complexity tasks, extended thinking provides clear
advantages, demonstrating the intended benefit of reasoning-augmented
models. However, in high-complexity tasks, both standard and reasoning
models exhibit complete accuracy collapse—a fundamental ceiling that
persists regardless of model capability or available computational
resources.

Particularly striking is their observation that reasoning models exhibit
*declining effort* at higher complexity levels despite adequate token
budgets. Rather than leveraging available computational resources more
intensively when problems become harder, models show inconsistent
reasoning patterns across similar problems. This suggests that current
reasoning mechanisms are not genuinely algorithmic but rather
approximate heuristics that fail when problems exceed certain complexity
thresholds.

For Black Box, these findings predict a non-linear relationship between
game difficulty (determined by atom configurations, number of atoms, and
observation ambiguity) and model performance. Simple configurations may
be solved adequately, but complex configurations requiring deep
inference chains should trigger the same "reasoning collapse" observed
across other puzzle domains.

## World Models and Mental Simulation

A "world model" refers to an internal representation that can be used to
simulate how states evolve over time. The question of whether LLMs
possess genuine world models—versus sophisticated pattern matching—is
actively debated.

@harig2025worldmodels investigated this question using mechanical
reasoning about pulley systems. Their findings are striking: while LLMs
showed some ability to estimate mechanical advantage, they appeared to
use a "pulley counting heuristic" rather than genuine simulation. The
best model (GPT-4o) achieved only 26.1% accuracy on exact predictions.
When problems were modified to be superficially similar but mechanically
different, performance collapsed, suggesting "abilities are often
fragile and dependent on superficial features rather than deep
understanding."

@wang2024bytesized directly tested whether LLMs can serve as text-based
world simulators. Using a benchmark of 76,369 state transitions from
text adventure games, they found that GPT-4's best single-step accuracy
was 59.9%. Because simulation errors accumulate, after 10 steps the
expected accuracy drops to less than 1% ($0.599^{10}$). They conclude
that "LLMs are not yet able to reliably act as text world simulators."

This literature suggests that LLMs may appear to reason about world
dynamics but are actually relying on heuristic shortcuts that fail when
problems require genuine simulation. For Black Box's "Predict"
mode—where the model must trace a ray's path given known atom
positions—we should expect systematic errors reflecting heuristic
approximation rather than accurate simulation.

## Abductive Reasoning and Hypothesis Generation

Abductive reasoning—inferring the most plausible explanation for
observations—is central to scientific discovery and diagnosis. A growing
body of research examines LLM capabilities in this domain.

@yang2025survey provides a comprehensive framework for understanding LLM
hypothesis discovery, organizing methods according to Peirce's reasoning
paradigm: abduction (hypothesis generation), deduction (hypothesis
application), and induction (hypothesis validation). They note that
"most existing work assessing LLM reasoning centers almost exclusively
on multi-step deductive tasks, leaving abduction and induction, the
engines of hypothesis discovery, largely unexplored."

@wang2025occam directly tests whether LLMs follow Occam's Razor—the
principle of preferring simpler explanations. Using a synthetic
benchmark (InAbHyD) designed to test inductive and abductive reasoning,
they find that "current LLMs can't generate high-quality hypotheses that
match human-level intelligence." LLMs frequently produce overly complex
explanations when simpler ones suffice, failing a fundamental principle
of scientific reasoning.

@liu2024incomplete characterizes the current state as "an incomplete
loop," arguing that while LLMs can perform some aspects of each
reasoning type, they fail to integrate abduction, deduction, and
induction into a coherent reasoning cycle. This fragmentation prevents
genuine hypothesis-driven inquiry.

@rodriguez2025analogical argues from a theoretical perspective that LLMs
face fundamental limitations in abductive reasoning: "(1) They cannot
work with counterfactuals, relying only on correlational datasets. (2)
They are unable to perform true abductive reasoning." While LLMs may
appear to generate hypotheses, this reflects parameter-controlled
variation rather than genuine inference to the best explanation.

For Black Box, the inverse problem—inferring atom locations from ray
observations—is quintessentially abductive. The literature predicts
systematic failures in generating parsimonious hypotheses and in
appropriately updating beliefs as evidence accumulates.

## Base Rate Neglect in Clinical Diagnosis

A particularly consequential manifestation of abductive reasoning
failure is base rate neglect—the tendency to overweight salient
observations while underweighting prior probabilities. @omar2025zebras
provide the first large-scale demonstration of this phenomenon in LLMs
applied to clinical diagnosis.

Using 300 synthetic clinical vignettes tested across 20 LLMs with
approximately 1.8 million responses, they found that models
systematically selected rare "zebra" diagnoses linked to salient but
non-decisive cues (such as travel history or animal exposure) more often
than epidemiologically correct diagnoses (49.8% vs 37.7%). This pattern
persisted even when models were explicitly asked to identify the "most
likely" diagnosis. Critically, prompts that explicitly invoked base
rates and epidemiology reduced but did not eliminate this saliency bias.

The implications are profound: LLMs that achieve high performance on
medical board examinations—demonstrating factual knowledge of disease
prevalence and diagnostic criteria—nevertheless fail to appropriately
weight this knowledge when salient distractors are present. As the
authors note, this represents "associative pattern matching rather than
stable probabilistic reasoning." Training on medical literature, which
overrepresents rare and dramatic conditions in case reports and teaching
materials, may systematically bias models toward zebra diagnoses.

For Black Box, this finding suggests that even when LLMs possess
accurate knowledge of ray physics, they may be systematically distracted
by salient features of observations (e.g., absorption events, unusual
deflection patterns) at the expense of systematic constraint tracking
and hypothesis evaluation. The base rate neglect documented in clinical
diagnosis may manifest as premature hypothesis commitment or failure to
consider the full space of configurations consistent with accumulated
evidence.

## Scientific Reasoning and Discovery

The limitations documented above converge on a concerning picture for
LLM deployment in scientific contexts.

@preprints2025scaling argues that LLM limitations in scientific
reasoning "stem not from insufficient data or compute, but from
architectural misalignment with the epistemic demands of science." They
propose that next-generation scientific LLMs must move beyond
token-level prediction to embrace structure-augmented architectures,
including integration with symbolic reasoning systems.

@liang2025hypothesis's survey of hypothesis generation identifies
persistent challenges: "Evaluating systems for scientific hypothesis
generation is a complex task." Key issues include ensuring novelty (not
just retrieving known hypotheses), grounding outputs in verifiable
knowledge, and enabling the model to articulate reasoning pathways.

The LLM-Modulo framework [@kambhampati2024llmmodulo;
@valmeekam2023planning] proposes a pragmatic response: accept that LLMs
cannot plan or reason reliably, but use them as components within hybrid
systems that include external verifiers and symbolic reasoners. Under
this view, the LLM's role is translation and suggestion, not autonomous
reasoning.

## Summary: Theoretical Predictions for Black Box

Based on the reviewed literature, we derive the following predictions:

1.  **Forward reasoning (Predict mode)** will show substantial errors,
    consistent with documented failures in spatial reasoning and world
    model simulation.

2.  **Inverse reasoning (Play mode)** will show even greater errors, as
    it compounds spatial reasoning with abductive inference and
    constraint satisfaction.

3.  **Prompt engineering** will provide modest improvements but will not
    overcome fundamental limitations.

4.  **Scaling** (larger models) will show improvement but not to
    human-level performance.

5.  **Error patterns** will reveal heuristic approximation rather than
    systematic reasoning—e.g., over-reliance on surface features,
    failure to track constraints, non-parsimonious hypotheses.

6.  **Tool augmentation** (e.g., giving the LLM a ray-tracing verifier)
    will help with verification but not with hypothesis generation, as
    the limitation is in search/inference, not just simulation.

7.  **Complexity-dependent collapse**: Following @shojaee2025illusion,
    performance will show a non-linear pattern—adequate performance on
    simple configurations but catastrophic failure on complex
    configurations requiring extended inference chains, regardless of
    available thinking budget.

8.  **Saliency-driven distraction**: Following @omar2025zebras, models
    will be systematically distracted by salient observations (e.g.,
    absorption events, dramatic deflections) and may anchor prematurely
    on hypotheses suggested by striking but non-decisive evidence, even
    when simpler explanations are more consistent with the full
    observation set.

# Methods

## The Black Box Game Implementation

We implemented Black Box as a web-based application using React,
allowing both human play and automated LLM interaction through the
Anthropic API. The implementation follows standard Black Box rules:

-   **Grid**: 8×8 cells, with 4 randomly placed atoms (non-adjacent
    placement enforced)
-   **Rays**: Fired from any of 32 edge positions (8 per side)
-   **Ray behavior**:
    -   Direct hit on atom: **Absorbed** (marked "H" for hit)
    -   Atom diagonally adjacent to entry cell: **Reflected** (marked
        "R")
    -   Atom diagonally adjacent to path: **Deflected** 90° away from
        atom
    -   Atoms on both sides of path: **Reflected** (reverses direction)
    -   Clear path: Ray exits at opposite edge

## Task Modes

### Predict Mode (Forward Reasoning)

In Predict mode, the LLM is shown: - The 8×8 grid with atom positions
marked - A ray entry point (e.g., "NORTH-4")

The LLM must predict the ray's outcome: exit position, absorption, or
reflection. This tests world model accuracy—can the LLM correctly
simulate ray physics?

### Play Mode (Inverse Reasoning)

In Play mode, the LLM: - Receives the game rules - Chooses which rays to
fire (up to 20) - Receives feedback after each ray (exit position,
absorption, or reflection) - Must guess all 4 atom positions

This tests the full reasoning pipeline: experiment design, observation
integration, constraint tracking, and abductive inference.

## Prompt Conditions

We developed two prompt styles with optional visualization augmentation:

### Baseline (Human-Equivalent)

Comprehensive instructions adapted from the traditional Emacs Black Box
rules, comparable to what a human player would receive. The prompt
includes:

-   Complete rules with ASCII diagram examples showing deflection,
    reflection, and absorption
-   Scoring information (rays cost points, missed atoms cost 5 points
    each)
-   JSON response format specification

Key excerpt:

```         
Your opponent has hidden 4 balls within this box. By shooting rays into
the box and observing where they emerge, it is possible to deduce the
positions of the hidden balls.

There are three possible outcomes for each ray you send into the box:
- DETOUR: The ray is deflected and emerges somewhere other than where
  you sent it in.
- REFLECTION (R): The ray is reflected and emerges in the same place
  it was sent in.
- HIT (H): The ray strikes a ball directly and is absorbed.
```

### Augmented

Extended instructions including:

-   Explicit coordinate system explanation
-   Detailed deflection geometry with worked examples
-   Strategy guidance (triangulation, multiple rays)
-   Common error warnings
-   Explicit note that single observations are typically ambiguous

### Visualization Options

Either prompt style can be combined with:

-   **Include Visualization**: Text-based board visualization showing
    current state
-   **VoT (Visualization of Thought) Options**:
    -   Grid State (Play only): Shows current board state in prompt
    -   Ray Trace: Shows ray path visualization
    -   Hypothesis (Play only): Shows marked hypothesis positions

## Models

We evaluated three Claude 4.5 model variants:

-   **Claude Haiku 4.5** (`claude-haiku-4-5-20251001`): Smallest,
    fastest
-   **Claude Sonnet 4.5** (`claude-sonnet-4-5-20250929`): Mid-tier
-   **Claude Opus 4.5** (`claude-opus-4-5-20251101`): Largest, most
    capable

All models were accessed through the Anthropic API with extended
thinking enabled to capture reasoning processes.

## Experimental Design

The experiment uses a factorial design with factors that differ slightly
between Predict and Play modes. Ten fixed atom configurations ensure
reproducibility across conditions.

### Predict Mode Factors

```{r}
#| label: predict-design-table
#| echo: false

predict_design <- tibble(
  Factor = c('Model', 'Prompt Style', 'Include Visualization', 'Extended Thinking', 'VoT Ray Trace', 'Atom Configuration'),
  Levels = c('Haiku 4.5, Sonnet 4.5, Opus 4.5',
             'Baseline, Augmented',
             'Yes, No',
             'Yes (with budget), No',
             'Yes, No',
             '10 fixed configurations'),
  Description = c('Model variant',
                  'Human-equivalent vs. detailed strategy guidance',
                  'Text board visualization in prompt',
                  'Enable extended thinking with token budget',
                  'Include ray trace visualization in prompt',
                  'Predetermined atom placements for reproducibility')
)

kable(predict_design, format = "pipe")
```

| Model      | Prompt   | Extended Thinking | VoT (None, A, B, or C)    |
|------------|----------|-------------------|---------------------------|
| Haiku 4.5  | baseline | No                | None                      |
| Sonnet 4.5 | baseline | No                | None                      |
| Opus 4.5   | baseline | No                | None                      |
| Haiku 4.5  | baseline | 10K               | None                      |
| Sonnet 4.5 | baseline | 10K               | None                      |
| Opus 4.5   | baseline | 10K               | None                      |
| Opus 4.5   | baseline | 10K               | None                      |
| Haiku 4.5  | baseline | 10K               | VoT A Grid State Tracking |
| Sonnet 4.5 | baseline | 10K               | VoT A Grid State Tracking |
| Opus 4.5   | baseline | 10K               | VoT A Grid State Tracking |

: Play Mode Experiment Log

### Play Mode Factors

Play mode includes all Predict mode factors plus additional factors for
hypothesis management:

```{r}
#| label: play-design-table
#| echo: false

play_design <- tibble(
  Factor = c('Model', 'Prompt Style', 'Include Visualization', 'Extended Thinking', 'Allow Hypotheses', 'VoT Grid State', 'VoT Ray Trace', 'VoT Hypothesis', 'Atom Configuration'),
  Levels = c('Haiku 4.5, Sonnet 4.5, Opus 4.5',
             'Baseline, Augmented',
             'Yes, No',
             'Yes (with budget), No',
             'Yes, No',
             'Yes, No',
             'Yes, No',
             'Yes, No',
             '10 fixed configurations'),
  Description = c('Model variant',
                  'Human-equivalent vs. detailed strategy guidance',
                  'Text board visualization in prompt',
                  'Enable extended thinking with token budget',
                  'Enable mark/unmark/check actions for hypothesis tracking',
                  'Show current grid state in prompt (VoT Option A)',
                  'Include ray trace visualization (VoT Option B)',
                  'Show hypothesized atom positions (VoT Option C)',
                  'Predetermined atom placements for reproducibility')
)

kable(play_design, format = "pipe")
```

### Atom Configurations

Ten fixed atom configurations are used for reproducible experiments:

```{r}
#| label: config-table
#| echo: false

configs <- tibble(
  Config = 1:10,
  Pattern = c('Spread', 'Cluster in corner', 'Diagonal', 'Edge-heavy', 'Central cluster',
              'L-shape', 'Corners', 'Asymmetric', 'Row cluster', 'Mixed'),
  Positions = c('(2,3), (3,6), (6,2), (7,7)',
                '(1,1), (1,3), (2,2), (5,6)',
                '(2,2), (4,4), (6,6), (8,8)',
                '(1,4), (4,8), (8,5), (5,1)',
                '(3,4), (4,3), (4,5), (5,4)',
                '(2,2), (2,3), (2,4), (4,2)',
                '(1,1), (1,8), (8,1), (8,8)',
                '(2,7), (3,2), (6,5), (7,3)',
                '(4,2), (4,4), (4,6), (4,8)',
                '(1,5), (3,3), (5,7), (8,2)')
)

kable(configs, format = "pipe")
```

### Predict Mode Procedure

For each experimental condition (model × prompt style × visualization ×
thinking × VoT):

1.  For each of the 10 fixed atom configurations:
    a.  For each of 32 possible ray entry points:
        -   Present atom positions and ray entry to LLM
        -   Record predicted outcome (exit position, absorption, or
            reflection)
        -   Compare to ground-truth ray trace
    b.  Compute accuracy metrics for this configuration
2.  Aggregate accuracy across configurations

### Play Mode Procedure

For each experimental condition (model × prompt style × visualization ×
thinking × hypotheses × VoT options):

1.  For each of the 10 fixed atom configurations:
    a.  Initialize game with hidden atom positions
    b.  LLM selects rays iteratively (max 20)
    c.  After each ray, provide outcome feedback
    d.  If hypotheses enabled, LLM can mark/unmark suspected positions
    e.  LLM submits final guess of 4 atom positions (via guess or check
        action)
    f.  Record: rays used, invalid moves, atoms correct, reasoning
        trace, thinking
2.  Aggregate performance metrics across configurations

## Dependent Variables

### Predict Mode

-   **Accuracy**: Proportion of rays with correct outcome prediction
-   **Error types**: Absorption/deflection/reflection confusion patterns
-   **Response time**: API response latency (ms)
-   **Token usage**: Input and output tokens per prediction

### Play Mode

-   **Atoms correct**: 0-4 atoms in correct positions
-   **Rays used**: Number of rays before guessing
-   **Invalid moves**: Attempts to fire into used positions
-   **Efficiency**: Atoms correct / rays used
-   **Score**: Game score (lower is better: ray costs + 5 × missed
    atoms)
-   **Response time**: Total API response time (ms)
-   **Token usage**: Total input and output tokens

### Process Measures (from Extended Thinking)

-   **Thinking traces**: Full extended thinking content when enabled
-   **Reasoning explanations**: Model's stated reasoning for each action
-   **Hypothesis tracking**: Marked positions over time (when enabled)
-   **Constraint tracking accuracy**: Quality of constraint maintenance
-   **Hypothesis revision patterns**: How beliefs change with new
    evidence

## Analysis Plan

### Primary Analyses

1.  **Predict Mode**: 3 (Model) × 3 (Prompt) ANOVA on accuracy
2.  **Play Mode**: Same design on atoms correct
3.  **Scaling analysis**: Linear trend across model sizes

### Qualitative Analysis

Systematic coding of extended thinking traces to identify: - Constraint
tracking errors - Spatial reasoning errors - Hypothesis generation
patterns - Response to contradictory evidence

# Results

## Predict Mode Results

```{r}
#| label: predict-results-placeholder
#| eval: false

# Placeholder for predict mode analysis
# Data will be loaded from experimental results

# predict_data <- jsonlite::fromJSON("predict_results.json")
#
# # Summary statistics
# predict_summary <- predict_data |>
#   group_by(model, prompt_condition) |>
#   summarise(
#     mean_correct = mean(correct),
#     sd_correct = sd(correct),
#     n = n(),
#     .groups = "drop"
#   ) |>
#   mutate(across(where(is.numeric), \(x) round(x, 3)))
#
# kable(predict_summary)
```

\[Results to be added after data collection\]

## Play Mode Results

```{r}
#| label: play-results-placeholder
#| eval: false

# Placeholder for play mode analysis
# Data will be loaded from experimental results

# play_data <- jsonlite::fromJSON("play_results.json")
#
# # Summary statistics
# play_summary <- play_data |>
#   group_by(model, prompt_condition) |>
#   summarise(
#     atoms_correct_mean = mean(atoms_correct),
#     atoms_correct_sd = sd(atoms_correct),
#     rays_used_mean = mean(rays_used),
#     rays_used_sd = sd(rays_used),
#     invalid_moves_mean = mean(invalid_moves),
#     invalid_moves_sd = sd(invalid_moves),
#     .groups = "drop"
#   ) |>
#   mutate(across(where(is.numeric), \(x) round(x, 2)))
#
# kable(play_summary)
```

\[Results to be added after data collection\]

## Error Analysis

```{r}
#| label: error-analysis-placeholder
#| eval: false

# Placeholder for error taxonomy
# Will categorize errors from extended thinking traces

# error_categories <- c(
#   "Constraint tracking failure",
#   "Spatial reasoning error",
#   "Non-parsimonious hypothesis",
#   "Failure to update on evidence",
#   "Experiment design inefficiency"
# )
```

\[Error analysis to be added after data collection\]

# Discussion

## Summary of Findings

\[To be completed after data collection\]

## Relation to Literature

### Constraint Satisfaction

\[Discuss how results relate to ZebraLogic, GenCP findings\]

### Spatial Reasoning

\[Discuss how results relate to PLUGH, spatial reasoning benchmarks\]

### World Models

\[Discuss how results relate to pulley system, text world simulator
findings\]

### Abductive Reasoning

\[Discuss how results relate to Occam's Razor, hypothesis generation
findings\]

## Implications

### For LLM Deployment in Diagnostic Domains

\[Discuss implications for medical diagnosis, scientific discovery\]

### For LLM-Modulo Architectures

\[Discuss whether tool augmentation addresses the limitations\]

### For Benchmark Development

\[Discuss Black Box as a diagnostic tool for reasoning assessment\]

## Limitations

1.  **Single game domain**: Results may not generalize to all reasoning
    tasks
2.  **Prompt sensitivity**: Different prompt formulations might yield
    different results
3.  **Model versions**: Rapid model updates may change performance
    characteristics
4.  **Sample size**: \[Discuss statistical power\]

## Future Directions

1.  **Extended model comparison**: GPT-4, Gemini, open-source models
2.  **Tool augmentation study**: Provide ray-tracing verification tool
3.  **Difficulty manipulation**: Vary grid size, atom count, atom
    arrangement
4.  **Human baseline**: Compare LLM performance to human players
5.  **Fine-tuning study**: Test whether task-specific training improves
    performance

# Conclusion

\[To be completed after data collection\]

# References

::: {#refs}
:::

# Appendix A: Prompt Texts {.appendix}

## Baseline Play Prompt

```         
You are playing Black Box, a game of hide and seek played on an 8 by 8 grid.

Your opponent has hidden 4 balls within this box. By shooting rays into the
box and observing where they emerge, it is possible to deduce the positions
of the hidden balls.

GRID: 8x8, rows 1-8 (top to bottom), columns 1-8 (left to right).
RAYS: Fire from edge positions - NORTH/SOUTH use columns 1-8, EAST/WEST use rows 1-8.

There are three possible outcomes for each ray you send into the box:

DETOUR: The ray is deflected and emerges somewhere other than where you sent
it in. Detours are denoted by matching pairs of numbers.

REFLECTION (R): The ray is reflected and emerges in the same place it was sent in.

HIT (H): The ray strikes a ball directly and is absorbed.

[ASCII diagram examples showing deflection, reflection, and hit patterns...]

SCORING:
Your goal is to minimize your score. Lower is better.
- Each ray entry point costs 1 point
- Each ray exit point costs 1 point (detours cost 2 total, reflections cost 1)
- Each missed atom costs 5 points

RULES:
- You cannot fire from positions already used as entry or exit points.
- Maximum 20 rays.

Respond with JSON only:
{"action": "fire", "side": "north|south|east|west", "position": 1-8, "reasoning": "..."}
{"action": "guess", "atoms": [[row,col], [row,col], [row,col], [row,col]], "reasoning": "..."}
```

## Baseline Predict Prompt

```         
Predict where a ray will exit in Black Box.

GRID: 8x8, rows 1-8 (top to bottom), columns 1-8 (left to right).
EDGES: NORTH/SOUTH use columns 1-8, EAST/WEST use rows 1-8.

As a ray approaches a ball it is deflected ninety degrees. Rays can be
deflected multiple times.

[ASCII diagram examples showing deflection, reflection, and hit patterns...]

Respond with JSON only:
{"exit_side": "north|south|east|west", "exit_position": 1-8, "reasoning": "..."}
OR for absorption: {"absorbed": true, "reasoning": "..."}
OR for reflection: {"reflected": true, "reasoning": "..."}
```

## Augmented Prompt

\[Full augmented prompt text to be included\]

## Hypothesis-Enabled Play Prompt Extension

When hypotheses are enabled, the response format is extended:

```         
Respond with JSON only:
{"action": "fire", "side": "north|south|east|west", "position": 1-8, "reasoning": "..."}
{"action": "mark", "row": 1-8, "col": 1-8, "reasoning": "..."} - mark suspected atom
{"action": "unmark", "row": 1-8, "col": 1-8, "reasoning": "..."} - remove marked position
{"action": "check", "reasoning": "..."} - submit answer (requires exactly 4 marked positions)
```

# Appendix B: Sample Game Traces {.appendix}

\[Representative game traces showing LLM reasoning\]

# Appendix C: Error Taxonomy {.appendix}

\[Detailed coding scheme for error analysis\]
